---
title: "DATA607 Project3 Part 2 — Extended EDA, Visualizations, and Thorough Analysis"
author: "Mehreen Gillani; Taha Malik; Paula Brown"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
params:
  data_dir: "."

---
# Rendering options
```{r setup2, include=FALSE} 
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6, dpi = 150
)
```
### Step 1. Import Libraries

```{r import libraries}
req_packages <- c("DBI", "RMySQL", "dplyr", "dbplyr", "knitr", "tidyr", "readr", "stringr", "tibble", "rmarkdown", "purrr", "lubridate", "here", "httr2", "ggplot2", "RCurl", "mice", "naniar", "gtExtras", "minqa", "patchwork", "forcats", "tools", "janitor", "skimr", "scales", "viridis", "corrplot", "broom", "pROC", "kableExtra")
for (pkg in req_packages) {
  if (!require(pkg, character.only = TRUE)) {
    message(paste("Installing package:", pkg))
    install.packages(pkg, dependencies = TRUE)
  } else {
    message(paste(pkg, " already installed."))
  }
  library(pkg, character.only = TRUE)
}
```

**Data Cleaning of Users dataset of Netflix - (Mehreen Gillani)**

### Step 2. Read users dataset and store it in a Dataframe

```{r read-csv}
users = read.csv('https://raw.githubusercontent.com/mehreengillani/DATA607_project3/refs/heads/main/users.csv')
colnames(users)
user <- users %>%
  distinct(user_id)
print(user)
```

### Step 3: Data Cleaning

####  3.1 Handling missing values
```{r identifymissingvalues}

# Identify missing values
miss_var_summary(users) %>%
  gt() %>%
  gt_theme_guardian() %>%
  tab_header(title="Missing data in users table")

#plot
gg_miss_var(users)
``` 

Find missing data pattern

```{r}
md.pattern(users)
```


By looking at the pattern we can see there are 21 rows where all 3 column values are missing.

We have missing data in 3 columns age, monthly_spend and household_size.
With missing percentages of 11.93%, 9.87%, and 15% in our Netflix users dataset, 
deletion would be a poor choice as we'd lose too much data. Here's the best strategy for handling this:



```{r}
# Get data types for specific columns
class(users$household_size)
class(users$monthly_spend)
```

Check density plot for handling missing data

```{r}
# Density plots show the shape better
par(mfrow = c(1, 3))

plot(density(users$age, na.rm = TRUE), main = "Age Density")
plot(density(users$monthly_spend, na.rm = TRUE), main = "Spend Density") 
plot(density(users$household_size, na.rm = TRUE), main = "Household Size Density")

par(mfrow = c(1, 1))
```

####  Perform multiple imputation
```{r}

imputed_data <- mice(users, 
                    m = 5,           # Create 5 complete datasets
                    method = 'pmm',  # Predictive Mean Matching
                    maxit = 10,      # Number of iterations
                    seed = 123)

# Check convergence (look for lines to converge)
plot(imputed_data)

# Create one completed dataset (or analyze all 5 and pool results)
users_complete <- complete(imputed_data, 1)

# Verify the imputation worked
sapply(users_complete, function(x) sum(is.na(x)))

```

Check density plot after imputation to see if our distribution is still the same 
```{r}
# Density plots after imputation
par(mfrow = c(1, 3))

plot(density(users_complete$age, na.rm = TRUE), main = "Age Density")
plot(density(users_complete$monthly_spend, na.rm = TRUE), main = "Spend Density") 
plot(density(users_complete$household_size, na.rm = TRUE), main = "Household Size Density")

par(mfrow = c(1, 1))
```

```{r}
# Set up plotting area to show all three
par(mfrow = c(1, 3))  # 1 row, 3 columns

# Create histograms
hist(users_complete$age, main = "Age Distribution", xlab = "Age", col = "lightblue")
hist(users_complete$monthly_spend, main = "Monthly Spend Distribution", xlab = "Spend", col = "lightgreen")
hist(users_complete$household_size, main = "Household Size Distribution", xlab = "Size", col = "lightcoral")

# Reset plotting area
par(mfrow = c(1, 1))
```
The density plot analysis reveals distinct distribution patterns across key variables. 
The age variable maintains its approximately normal distribution following imputation, indicating the preservation of natural demographic patterns. <br>
In contrast, both monthly spending and household size exhibit significant right-skewness.  <br>
The monthly spend distribution features a concentration of users at lower price points with a long tail representing premium subscribers,  <br>
while household size distribution is dominated by single-person accounts with progressively fewer multi-person households.  <br>
Both monthly_spend and household_size distribution didn't change after performing multiple imputation. <br>


```{r check_missing_values}
#check if there is still any missing data
miss_var_summary(users_complete)
```


#### Step 3.2: Removing Duplicates

```{r checkduplicates}
# Identify total duplicates
sum(duplicated(users_complete))
duplicated_percent <- (sum(duplicated(users_complete)) / nrow(users_complete) * 100)
print(duplicated_percent)

```

The percentage of duplicates is 2.34 so we will delete duplicated rows

```{r remove-duplicates}

# Remove ALL duplicate rows (keeping first occurrence)
users_unique <- distinct(users_complete)

# Verify removal
print(paste("Before:", nrow(users_complete), "rows"))
print(paste("After:", nrow(users_unique), "rows"))
print(paste("Removed:", nrow(users_complete) - nrow(users_unique), "duplicate rows"))
```

####  Step: 3.3 Data Type Conversion

```{r check _data_ types}
# Basic structure
str(users_unique)
```
```{r datatype conversion}
users_improved <- users_unique %>%
  mutate(
    # 1. Convert dates from character to proper date format
    subscription_start_date = as.Date(subscription_start_date),
    created_at = as.POSIXct(created_at),
    
    # 2. Convert boolean from character to logical
    is_active = case_when(
      is_active == "True" ~ TRUE,
      is_active == "False" ~ FALSE,
      TRUE ~ NA
    ),
    
    # 3. Convert categorical variables to factors
    gender = factor(gender),
    country = factor(country),
    state_province = factor(state_province),
    city = factor(city),
    subscription_plan = factor(subscription_plan),
    primary_device = factor(primary_device),
    
    # 4. Ensure numeric columns are proper integers where appropriate
    age = as.integer(age),
    household_size = as.integer(round(household_size)),  # Household size should be whole numbers
    
    # 5. Keep user_id and email as character (correct)
    # monthly_spend as numeric is correct (can have decimals)
  )
```

We have converted created_at to date format, is_active to logical and categorical values to factors

```{r}
#check new structure 
str(users_improved)
```
```{r}
# Check for any remaining issues
summary(users_improved)

# Verify no data loss occurred
stopifnot(nrow(users_unique) == nrow(users_improved))

# Check factor levels
sapply(users_improved[c("gender", "subscription_plan", "primary_device")], levels)
```
<b>subscription plan factors need to be revised </b>

```{r proper ordering for subscription plan}
users_improved <- users_improved %>%
  mutate(
    subscription_plan = factor(subscription_plan, 
                              levels = c("Basic", "Standard", "Premium", "Premium+"),
                              ordered = TRUE)
  )
```

I have noticed there are negative values in age variable  lets deal with that problem. age cannot be negative.

```{r}
min(users_improved$age)

# Count how many negative values exist
negative_age_count <- sum(users_improved$age < 0, na.rm = TRUE)
cat("Number of negative age values:", negative_age_count, "\n")

# Calculate median age from valid values only
median_age <- median(users_improved$age[users_improved$age >= 0], na.rm = TRUE)

# Replace negative values with median
users_improved <- users_improved %>%
  mutate(age = ifelse(age < 0, median_age, age))

# Verify
summary(users_improved$age)
```
I have noticed an empty strings ("") in the gender column, which should be handled. Here are the best approaches:

```{r}
users_improved <- users_improved %>%
  mutate(
    # Step 1: Convert empty strings to "Unknown"
    gender = fct_recode(gender, 'Unknown' = ""),
    
    # Step 2: Consolidate categories
    gender = fct_collapse(gender,
                         "Other" = c("Other", "Prefer not to say")
    ),
    
    # Step 3: Relevel in the correct order
    gender = fct_relevel(gender, "Female", "Male", "Other", "Unknown")
  )

# Verify
cat("Gender levels in order:\n")
levels(users_improved$gender)
cat("Gender distribution:\n")
table(users_improved$gender)
```
Instead of empty string I have filled it with unknown 

####  Step 3.4. Handling Outliers

```{r fine numericcolumns}
# Find numeric columns
numeric_cols <- names(users_improved)[sapply(users_improved, is.numeric)]

cat("Numeric columns:\n")
print(numeric_cols)
```


```{r handling-outliers}
# Cap outliers instead of removing them (keeps all rows)
users_clean <- users_improved %>%
  mutate(across(where(is.numeric), ~ {
    if(all(is.na(.))) return(.)
    Q1 <- quantile(., 0.25, na.rm = TRUE)
    Q3 <- quantile(., 0.75, na.rm = TRUE)
    IQR <- Q3 - Q1
    lower <- Q1 - 1.5 * IQR
    upper <- Q3 + 1.5 * IQR
    case_when(
      . < lower ~ lower,
      . > upper ~ upper,
      TRUE ~ .
    )
  }))
```

BoxPlot before Outlier handling 

```{r before-outlier-detection}

boxplot(users_improved$age, main = "Age Outliers", col = "lightblue")
boxplot(users_improved$monthly_spend, main = "Monthly Spend Outliers", col = "lightgreen")
boxplot(users_improved$household_size, main = "Household Size Outliers", col = "lightcoral")

# Reset plot layout
par(mfrow = c(1, 1))
```

BoxPlot after outlier handling with cap method

```{r after-outlier-detection}
boxplot(users_clean$age, main = "Age Outliers", col = "lightblue")
boxplot(users_clean$monthly_spend, main = "Monthly Spend Outliers", col = "lightgreen")
boxplot(users_clean$household_size, main = "Household Size Outliers", col = "lightcoral")

# Reset plot layout
par(mfrow = c(1, 1))
```

Check how many outliers we had before placing cap

```{r check number of outlera}
# Check how many outliers we have before placing cap
for(col in numeric_cols) {
  data <- users_improved[[col]]
  if(!all(is.na(data))) {
    Q1 <- quantile(data, 0.25, na.rm = TRUE)
    Q3 <- quantile(data, 0.75, na.rm = TRUE)
    IQR <- Q3 - Q1
    outliers <- sum(data < (Q1 - 1.5*IQR) | data > (Q3 + 1.5*IQR), na.rm = TRUE)
    cat(col, ":", outliers, "outliers\n")
  }
}
```


####  3.5 String Cleaning


```{r find-textCols_string_cleaning}
# Find all text columns
text_cols <- names(users_clean)[sapply(users_clean, function(x) is.character(x) | is.factor(x))]

cat("Text columns found:\n")
print(text_cols)
```

Text cleaning: convert to lower case, convert empty string, remove whitespace
```{r txt_cleaning_p2}
# Perform comprehensive text cleaning
users_clean <- users_clean %>%
  mutate(across(all_of(text_cols), ~ {
    if(is.character(.)) {
      # Remove leading/trailing whitespace
      cleaned <- trimws(.)
      # Convert to lowercase for standardization
      cleaned <- tolower(cleaned)
      # Remove multiple spaces
      cleaned <- gsub("\\s+", " ", cleaned)
      # Convert empty strings to NA
      cleaned <- ifelse(cleaned == "", NA, cleaned)
      return(cleaned)
    } else {
      return(.)
    }
  }))
```


####  3.6 Data Validation

```{r}
# Check for impossible values
invalid_ages <- users_clean$age < 0 | users_clean$age > 120
invalid_dates <- users_clean$date > Sys.Date()

# Fix invalid values
users_clean$age[users_clean$age < 0 | users_clean$age > 120] <- NA

# Check consistency
inconsistent <- users_clean$end_date < users_clean$start_date

```

Verify all issues are resolved

```{r checkdata quality}

cat("=== FINAL DATA QUALITY CHECK ===\n")
cat("Ages < 0:", sum(users_clean$age < 0, na.rm = TRUE), "\n")
cat("Ages > 120:", sum(users_clean$age > 120, na.rm = TRUE), "\n")
cat("Negative monthly spend:", sum(users_clean$monthly_spend < 0, na.rm = TRUE), "\n")
cat("Household size < 1:", sum(users_clean$household_size < 1, na.rm = TRUE), "\n")
cat("Household size > 20:", sum(users_clean$household_size > 20, na.rm = TRUE), "\n")
cat("Empty gender:", sum(users_clean$gender == "" | is.na(users_clean$gender), na.rm = TRUE), "\n")

# Summary of cleaned data
cat("\n=== CLEANED DATA SUMMARY ===\n")
summary(users_clean[c("age", "monthly_spend", "household_size", "gender")])

```


This validation step is crucial before any analysis to ensure your results are reliable and meaningful!

Write cleaned data to another csv file:

```{r}
# Write to CSV file
write.csv(users_clean, "users_clean.csv", row.names = FALSE)

cat("Cleaned data saved as 'users_clean.csv'\n")
cat("File location:", getwd(), "\n")

```

```{r checknewfile}
users_clean_data <- read_csv('users_clean.csv')
str(users_clean_data)

```

**Data Cleaning of Movies dataset of Netflix - (Mehreen Gillani)**

### Step 2. Read movies dataset and store it in a Dataframe

```{r read}
movie_url = 'https://raw.githubusercontent.com/mehreengillani/DATA607_project3/refs/heads/main/movies.csv'
movies = read.csv(movie_url)
colnames(movies)

#printing unique movie id's
movie <- movies %>%
  distinct(movie_id)
print(movie)
unique(movies$movie_id)
#printing unique movie titles
unique(movies$title)
#data structure
str(movies)
```

### Step 3: Data Cleaning

####  3.1 Handling missing values
```{r identify-missing-values}

# Identify missing values
miss_var_summary(movies) %>%
  gt() %>%
  gt_theme_guardian() %>%
  tab_header(title="Missing data in movies table")

#plot
gg_miss_var(movies)
``` 

Immediate Assessment

High missingness (>60%): number_of_seasons, number_of_episodes, box_office_revenue, production_budget
Moderate missingness (14.4%): imdb_rating



Find missing data pattern

```{r}
md.pattern(movies)
```


By looking at the pattern we can see there are 42 rows where all 5 column values are missing.



Setting number_of_seasons and number_of_episodes to 0 for movies, since movies don't have seasons or episodes
If it is a TV Show with missing seasons/episodes?
  * Use median of other TV shows, or 1 season/12 episodes as default
If it is missing for any other reason?
  * Set to 1 season/12 episodes (catch-all)
  
For budgeting and revenue:
  * Modeling: Can include budget_missing and revenue_missing as a predictor to account for data quality
  * 0 budget/revenue clearly indicates "data not available" rather than "actually zero"

```{r}
# Final comprehensive missing data handling


movies_handle_missing <- movies %>%
  mutate(
    # Handle number_of_seasons
    number_of_seasons = case_when(
      content_type == "Movie" & is.na(number_of_seasons) ~ 0,
      content_type == "TV Show" & is.na(number_of_seasons) ~ 
        coalesce(median(number_of_seasons[content_type == "TV Show"], na.rm = TRUE), 1),
      is.na(number_of_seasons) ~ 1,  # Catch any other missing
      TRUE ~ number_of_seasons
    ),
    # Handle number_of_episodes
    number_of_episodes = case_when(
      content_type == "Movie" & is.na(number_of_episodes) ~ 0,
      content_type == "TV Show" & is.na(number_of_episodes) ~ 
        coalesce(median(number_of_episodes[content_type == "TV Show"], na.rm = TRUE), 12),
      is.na(number_of_episodes) ~ 12,  # Catch any other missing
      TRUE ~ number_of_episodes
    )
  ) %>%
  mutate(
    budget_missing = as.integer(is.na(production_budget)),
    revenue_missing = as.integer(is.na(box_office_revenue)),
    production_budget = ifelse(is.na(production_budget), 0, production_budget),
    box_office_revenue = ifelse(is.na(box_office_revenue), 0, box_office_revenue)
  )

# verification
cat("=== FINAL MISSINGNESS CHECK ===\n")
print(miss_var_summary(movies_handle_missing))


```

####  Perform multiple imputation
```{r}
# Multiple Imputation on imdb_rating


# Create a clean dataset first
movies_impute <- movies_handle_missing %>%
  select(imdb_rating, duration_minutes, rating) %>%
  filter(complete.cases(duration_minutes, rating))  # Remove rows where predictors are missing

# Check the cleaned dataset
cat("Rows after removing missing predictors:", nrow(movies_impute), "\n")
cat("Missing imdb_rating in cleaned data:", sum(is.na(movies_impute$imdb_rating)), "\n")

# Now try imputation on the clean dataset
if(nrow(movies_impute) > 0 && sum(is.na(movies_impute$imdb_rating)) > 0) {
  imputed_data <- mice(movies_impute,
                      m = 3,
                      maxit = 5,
                      method = "pmm",
                      printFlag = TRUE,
                      seed = 123)
  movies_imputed_small <- complete(imputed_data, 1)
  
  # Merge back with original data to get all columns
  movies_complete <- movies_handle_missing %>%
    # Remove rows that were used in imputation (they'll be replaced)
    filter(!(complete.cases(duration_minutes, rating))) %>%
    # Bind with the imputed data
    bind_rows(
      movies_handle_missing %>%
        filter(complete.cases(duration_minutes, rating)) %>%
        select(-imdb_rating) %>%  # Remove original imdb_rating
        bind_cols(imdb_rating = movies_imputed_small$imdb_rating)  # Add imputed values
    )
  
  print(summary(movies_complete$imdb_rating))
  cat("Total rows in final dataset:", nrow(movies_complete), "\n")
  cat("Columns in final dataset:", names(movies_complete), "\n")
  
} else {
  cat("No data to impute after cleaning\n")
}
```

Check density plot after imputation to see if our distribution is still the same 
```{r}
# Density plots of imdb_ratings before and after imputation
par(mfrow = c(1, 2))
plot(density(movies$imdb_rating, na.rm = TRUE), main = "imdb_rating Density \n before multiple imputation")
plot(density(movies_complete$imdb_rating, na.rm = TRUE), main = "imdb_rating \n Density after imputation")


par(mfrow = c(1, 1))
```

Distribution is overall the same but there are some small differences in the density plot shape. 
Lets Calculate before/after statistics

```{r}

# Calculate before/after statistics
before <- movies_handle_missing$imdb_rating[!is.na(movies_handle_missing$imdb_rating)]
after <- movies_complete$imdb_rating

# Basic statistics comparison
stats_before <- c(
  mean = mean(before),
  sd = sd(before),
  median = median(before),
  min = min(before),
  max = max(before),
  n = length(before)
)

stats_after <- c(
  mean = mean(after),
  sd = sd(after),
  median = median(after),
  min = min(after),
  max = max(after),
  n = length(after)
)

comparison <- data.frame(Before = stats_before, After = stats_after, Change = stats_after - stats_before, percentage = ((stats_after - stats_before)/stats_before) * 100)
print(round(comparison, 3))
```


Interpretation of Results

✅ GOOD SIGNS:

Minimal mean change: -3.03% (from 6.269 to 6.078) <br>
Well within acceptable range (<5%) <br>

Small SD increase: +5.09% (from 1.809 to 1.901)
Within acceptable range (<15%)

Range preserved: Min (0.5) and Max (10.0) unchanged

Median stability: Small change from 6.4 to 6.3



```{r check-missing-values}
#check if there is still any missing data
miss_var_summary(movies_complete)
```


#### Step 3.2: Removing Duplicates

```{r checkdupes}
# Identify total duplicates
sum(duplicated(movies_complete))
duplicated_percent <- (sum(duplicated(movies_complete)) / nrow(movies_complete) * 100)
print(duplicated_percent)

```

The percentage of duplicates is 3.36 so we will delete duplicated rows

```{r remove_duplicates}

# Remove ALL duplicate rows (keeping first occurrence)
movies_unique <- distinct(movies_complete)

# Verify removal
print(paste("Before:", nrow(movies_complete), "rows"))
print(paste("After:", nrow(movies_unique), "rows"))
print(paste("Removed:", nrow(movies_complete) - nrow(movies_unique), "duplicate rows"))
```

####  Step: 3.3 Data Type Conversion

```{r check data-types}
# Basic structure
str(movies_unique)
```

```{r data-type conversion}


rating_order <- c("TV-Y", "TV-Y7", "G", "TV-G", "PG", "TV-PG", "PG-13", "TV-14", "R", "NC-17", "TV-MA")
movies_improved <- movies_unique %>%
  mutate(
    # 1. Convert logical strings to actual logical values
    is_netflix_original = as.logical(is_netflix_original),
    content_warning = as.logical(content_warning),
    
    # 2. Convert date strings to proper Date type
    added_to_platform = as.Date(added_to_platform),
    
    # 3. Convert categorical variables to factors
    content_type = as.factor(content_type),
    genre_primary = as.factor(genre_primary),
    genre_secondary = as.factor(genre_secondary),
    rating = factor(rating, 
                   levels = rating_order,
                   ordered = TRUE),
    language = as.factor(language),
    country_of_origin = as.factor(country_of_origin),
    
    # 4. Keep binary indicators as integers (they're fine as-is)
    # budget_missing and revenue_missing are correctly typed as integers
    
    # 5. Ensure numeric types are correct (they already are)
    # imdb_rating, production_budget, box_office_revenue, etc. are correctly numeric
  )
  
```

We have converted added_to_platform to date format, is_netflix_original and content_warning to logical and
categorical values to factors:
content_type, genre_primary, genre_secondary, rating, language, country_of_origin

```{r}
#check new structure 
str(movies_improved)
```

####  Step 3.4. Handling Outliers

```{r fine numeric columns}
# Find numeric columns
numeric_cols <- names(movies_improved)[sapply(movies_improved, is.numeric)]

cat("Numeric columns:\n")
print(numeric_cols)

```


```{r handling outliers}


# Create a copy for outlier treatment
movies_outlier_treated <- movies_improved

# Define outlier treatment for each column
outlier_summary <- list()

for(col in numeric_cols) {
  data <- movies_outlier_treated[[col]]
  
  if(!all(is.na(data))) {
    cat("\n=== Treating outliers in:", col, "===\n")
    
    # Store original stats
    original_stats <- c(
      min = min(data, na.rm = TRUE),
      max = max(data, na.rm = TRUE),
      mean = mean(data, na.rm = TRUE)
    )
    
    # Column-specific outlier treatment
    if(col == "release_year") {
      # For years: use reasonable bounds (e.g., 1900-current year+5)
      lower_cap <- 1900
      upper_cap <- as.numeric(format(Sys.Date(), "%Y")) + 5
      treated_data <- ifelse(data < lower_cap, lower_cap, 
                            ifelse(data > upper_cap, upper_cap, data))
      
    } else if(col == "duration_minutes") {
      # For duration: reasonable movie/TV show lengths
      lower_cap <- 1       # Minimum 1 minute
      upper_cap <- quantile(data, 0.99, na.rm = TRUE)  # Cap at 99th percentile
      treated_data <- ifelse(data < lower_cap, lower_cap, 
                            ifelse(data > upper_cap, upper_cap, data))
      
    } else if(col %in% c("imdb_rating")) {
      # For ratings: use actual scale bounds (0.5-10)
      lower_cap <- 0.5
      upper_cap <- 10.0
      treated_data <- ifelse(data < lower_cap, lower_cap, 
                            ifelse(data > upper_cap, upper_cap, data))
      
    } else if(col %in% c("production_budget", "box_office_revenue")) {
      # For financial data: 0 lower bound, 99.5% upper cap
      lower_cap <- 0
      upper_cap <- quantile(data, 0.95, na.rm = TRUE)
      treated_data <- ifelse(data < lower_cap, lower_cap, 
                            ifelse(data > upper_cap, upper_cap, data))
      
    } else if(col %in% c("number_of_seasons", "number_of_episodes")) {
      # For seasons/episodes: 0 lower bound, robust upper cap
      lower_cap <- 0
      upper_cap <- quantile(data, 0.99, na.rm = TRUE)
      treated_data <- ifelse(data < lower_cap, lower_cap, 
                            ifelse(data > upper_cap, upper_cap, data))
      
    } else if(col %in% c("budget_missing")) {
      # For binary indicators: no treatment needed (should be 0/1)
      treated_data <- data
      cat("  No treatment needed for binary indicator\n")
      next
    }
    
    # Apply the treatment
    movies_outlier_treated[[col]] <- treated_data
    
    # Calculate treatment impact
    new_stats <- c(
      min = min(treated_data, na.rm = TRUE),
      max = max(treated_data, na.rm = TRUE)
    )
    
    capped_low <- sum(data < lower_cap, na.rm = TRUE)
    capped_high <- sum(data > upper_cap, na.rm = TRUE)
    
    cat("  Lower cap:", lower_cap, "| Upper cap:", round(upper_cap, 2), "\n")
    cat("  Values capped - Low:", capped_low, "| High:", capped_high, "\n")
    cat("  Original range: [", original_stats["min"], ", ", original_stats["max"], "]\n", sep = "")
    cat("  New range: [", new_stats["min"], ", ", new_stats["max"], "]\n", sep = "")
    
    # Store summary
    outlier_summary[[col]] <- data.frame(
      variable = col,
      lower_cap = lower_cap,
      upper_cap = upper_cap,
      capped_low = capped_low,
      capped_high = capped_high,
      original_min = original_stats["min"],
      original_max = original_stats["max"],
      new_min = new_stats["min"],
      new_max = new_stats["max"]
    )
  }
}
```

BoxPlot before Outlier handling 

```{r before outlier detection}

boxplot(movies_improved$imdb_rating, main = "imdb_rating Outliers", col = "lightblue")
boxplot(movies_improved$production_budget, main = "production_budget Spend Outliers", col = "lightgreen")
boxplot(movies_improved$box_office_revenue, main = "box_office_revenue Size Outliers", col = "lightcoral")

# Reset plot layout
par(mfrow = c(1, 1))
```
```{r}
# Create comparison plots for key variables
create_comparison_plot <- function(col) {
  p1 <- ggplot(movies_improved, aes(x = .data[[col]])) +
    geom_histogram(fill = "red", alpha = 0.6, bins = 30) +
    ggtitle(paste("Original", col)) +
    theme_minimal()
  
  p2 <- ggplot(movies_outlier_treated, aes(x = .data[[col]])) +
    geom_histogram(fill = "blue", alpha = 0.6, bins = 30) +
    ggtitle(paste("Treated", col)) +
    theme_minimal()
  
  p1 + p2
}

# Plot comparisons for key columns
key_columns <- c("production_budget", "box_office_revenue", "duration_minutes", "number_of_seasons")
for(col in key_columns) {
  print(create_comparison_plot(col))
}

```

BoxPlot after outlier handling with cap method

```{r after outlier detection}

boxplot(movies_outlier_treated$imdb_rating, main = "imdb_rating Outliers", col = "lightblue")
boxplot(movies_outlier_treated$production_budget, main = "production_budget Spend Outliers", col = "lightgreen")
boxplot(movies_outlier_treated$box_office_revenue, main = "box_office_revenue Size Outliers", col = "lightcoral")
boxplot(movies_outlier_treated$duration_minutes, main = "duration_minutes Spend Outliers", col = "lightgreen")
boxplot(movies_outlier_treated$number_of_episodes, main = "number_of_episodes Size Outliers", col = "lightcoral")

# Reset plot layout
par(mfrow = c(1, 1))
```



```{r check number of outlier}
# Check how many outliers we have before placing cap
for(col in numeric_cols) {
  data <- movies_improved[[col]]
  if(!all(is.na(data))) {
    Q1 <- quantile(data, 0.25, na.rm = TRUE)
    Q3 <- quantile(data, 0.75, na.rm = TRUE)
    IQR <- Q3 - Q1
    outliers <- sum(data < (Q1 - 1.5*IQR) | data > (Q3 + 1.5*IQR), na.rm = TRUE)
    cat(col, ":", outliers, "outliers\n")
  }
}
```


####  3.5 String Cleaning


```{r find_textCols_string_cleaning}
# Find all text columns
text_cols <- names(movies_outlier_treated)[sapply(movies_outlier_treated, function(x) is.character(x) | is.factor(x))]

cat("Text columns found:\n")
print(text_cols)
```

Text cleaning: convert to lower case, convert empty string, remove whitespace
```{r txtcleaning_p2}
# Perform comprehensive text cleaning
movies_clean <- movies_outlier_treated %>%
  mutate(across(all_of(text_cols), ~ {
    if(is.character(.)) {
      # Remove leading/trailing whitespace
      cleaned <- trimws(.)
      # Convert to lowercase for standardization
      cleaned <- tolower(cleaned)
      # Remove multiple spaces
      cleaned <- gsub("\\s+", " ", cleaned)
      # Convert empty strings to NA
      cleaned <- ifelse(cleaned == "", NA, cleaned)
      return(cleaned)
    } else {
      return(.)
    }
  }))
```
```{r column names}
colnames(movies_clean)
```


####  3.6 Data Validation

```{r Validate categorical variables}
# Validate categorical variables
cat("\n=== CATEGORICAL VALUE VALIDATION ===\n")

categorical_vars <- c("content_type", "genre_primary", "rating", "language", "country_of_origin")

for(var in categorical_vars) {
  if(var %in% names(movies_clean)) {
    unique_vals <- unique(movies_clean[[var]])
    cat(var, ":", length(unique_vals), "unique values\n")
    
    # Show top 10 most frequent values
    if(length(unique_vals) > 0) {
      freq_table <- table(movies_clean[[var]])
      cat("  Top values:", names(head(sort(freq_table, decreasing = TRUE), 5)), "\n")
    }
    
    # Check for empty strings
    if(is.character(movies_clean[[var]])) {
      empty_count <- sum(movies_clean[[var]] == "" | is.na(movies_clean[[var]]), na.rm = TRUE)
      cat("  Empty/NA values:", empty_count, "\n")
    }
  }
}

```
```{r}
# Simple quality check without all the details
simple_quality_check <- function(movies_clean) {
  total_rows <- nrow(movies_clean)
  
  # Basic metrics
  completeness <- mean(complete.cases(movies_clean)) * 100
  duplicates <- sum(duplicated(movies_clean))
  uniqueness <- (1 - duplicates/total_rows) * 100
  
  # Simple quality score
  quality_score <- mean(c(completeness, uniqueness))
  
  cat("Simple Quality Assessment:\n")
  cat("Completeness:", round(completeness, 1), "%\n")
  cat("Uniqueness:", round(uniqueness, 1), "%\n") 
  cat("Overall Score:", round(quality_score, 1), "%\n")
  
  return(quality_score)
}

# Run simple version
simple_quality_check(movies_clean)

```
```{r}
# Complete data quality validation function
validate_data_quality <- function(df) {
  cat("=== COMPREHENSIVE DATA QUALITY VALIDATION ===\n")
  
  # Initialize all variables within the function
  total_rows <- nrow(df)
  total_cols <- ncol(df)
  
  # 1. Basic Structure Validation
  cat("\n1. BASIC STRUCTURE VALIDATION\n")
  cat("Total rows:", total_rows, "\n")
  cat("Total columns:", total_cols, "\n")
  cat("Complete cases:", sum(complete.cases(df)), "(", 
      round(mean(complete.cases(df)) * 100, 1), "%)\n")
  
  duplicates <- sum(duplicated(df))
  cat("Duplicate rows:", duplicates, "\n")
  
  # 2. Data Type Validation
  cat("\n2. DATA TYPE VALIDATION\n")
  data_types <- sapply(df, class)
  print(data_types)
  
  # 3. Value Range Validation
  cat("\n3. VALUE RANGE VALIDATION\n")
  
  # Define range checks within the function
  range_checks <- list(
    release_year = c(min = 1900, max = as.numeric(format(Sys.Date(), "%Y")) + 1),
    duration_minutes = c(min = 1, max = 500),
    imdb_rating = c(min = 0.5, max = 10.0),
    production_budget = c(min = 0, max = NA),
    box_office_revenue = c(min = 0, max = NA),
    number_of_seasons = c(min = 0, max = 50),
    number_of_episodes = c(min = 0, max = 500)
  )
  
  range_violations_total <- 0
  total_possible_checks <- 0
  
  for(var in names(range_checks)) {
    if(var %in% names(df)) {
      data <- df[[var]]
      range <- range_checks[[var]]
      
      violations_min <- sum(data < range["min"], na.rm = TRUE)
      violations_max <- ifelse(is.na(range["max"]), 0, sum(data > range["max"], na.rm = TRUE))
      total_violations <- violations_min + violations_max
      
      range_violations_total <- range_violations_total + total_violations
      total_possible_checks <- total_possible_checks + total_rows
      
      cat(var, ":\n")
      cat("  Range violations:", total_violations, "\n")
      cat("  Actual range: [", min(data, na.rm = TRUE), ", ", max(data, na.rm = TRUE), "]\n", sep = "")
    }
  }
  
  # 4. Business Logic Validation
  cat("\n4. BUSINESS LOGIC VALIDATION\n")
  logic_checks <- data.frame(
    Check = character(),
    Violations = integer(),
    stringsAsFactors = FALSE
  )
  
  # Add business logic checks
  if("content_type" %in% names(df) && "number_of_seasons" %in% names(df)) {
    movies_with_seasons <- sum(df$content_type == "Movie" & df$number_of_seasons > 0, na.rm = TRUE)
    logic_checks <- rbind(logic_checks, data.frame(
      Check = "Movies with seasons > 0",
      Violations = movies_with_seasons
    ))
  }
  
  if("content_type" %in% names(df) && "number_of_episodes" %in% names(df)) {
    movies_with_episodes <- sum(df$content_type == "Movie" & df$number_of_episodes > 0, na.rm = TRUE)
    logic_checks <- rbind(logic_checks, data.frame(
      Check = "Movies with episodes > 0",
      Violations = movies_with_episodes
    ))
  }
  
  if("production_budget" %in% names(df) && "box_office_revenue" %in% names(df)) {
    budget_exceeds_revenue <- sum(df$production_budget > df$box_office_revenue, na.rm = TRUE)
    logic_checks <- rbind(logic_checks, data.frame(
      Check = "Budget exceeds revenue",
      Violations = budget_exceeds_revenue
    ))
  }
  
  print(logic_checks)
  total_logic_violations <- sum(logic_checks$Violations)
  
  # 5. Missing Data Validation
  cat("\n5. MISSING DATA VALIDATION\n")
  missing_summary <- sapply(df, function(x) sum(is.na(x)))
  missing_df <- data.frame(
    Variable = names(missing_summary),
    Missing_Count = missing_summary,
    Missing_Percent = round(missing_summary / total_rows * 100, 2)
  ) %>% arrange(desc(Missing_Percent))
  
  print(head(missing_df, 10))  # Show top 10 most missing
  
  # 6. Calculate Quality Score
  cat("\n6. DATA QUALITY SCORE\n")
  
  quality_metrics <- c(
    completeness = mean(complete.cases(df)),
    uniqueness = ifelse(total_rows > 0, 1 - (duplicates / total_rows), 1),
    valid_ranges = ifelse(total_possible_checks > 0, 
                          1 - (range_violations_total / total_possible_checks), 1),
    business_rules = ifelse(total_rows > 0, 
                            1 - (total_logic_violations / (total_rows * nrow(logic_checks))), 1)
  )
  
  # Handle any NaN values
  quality_metrics <- sapply(quality_metrics, function(x) ifelse(is.nan(x), 1, x))
  
  quality_score <- mean(quality_metrics) * 100
  
  # Display quality metrics
  for(i in seq_along(quality_metrics)) {
    cat(names(quality_metrics)[i], ":", round(quality_metrics[i] * 100, 1), "%\n")
  }
  
  cat("\nOVERALL DATA QUALITY SCORE:", round(quality_score, 1), "%\n")
  
  # Quality rating
  if(quality_score >= 95) {
    rating <- "EXCELLENT"
    emoji <- "🎯"
  } else if(quality_score >= 85) {
    rating <- "VERY GOOD" 
    emoji <- "✅"
  } else if(quality_score >= 75) {
    rating <- "GOOD"
    emoji <- "✓"
  } else if(quality_score >= 65) {
    rating <- "ACCEPTABLE"
    emoji <- "⚠️"
  } else {
    rating <- "NEEDS IMPROVEMENT"
    emoji <- "❌"
  }
  
  cat(emoji, " DATA QUALITY: ", rating, "\n", sep = "")
  
  # Return summary
  return(list(
    quality_score = quality_score,
    rating = rating,
    duplicates = duplicates,
    range_violations = range_violations_total,
    logic_violations = total_logic_violations
  ))
}

# Run the validation
quality_results <- validate_data_quality(movies_clean)

```

This validation step is crucial before any analysis to ensure your results are reliable and meaningful!

Write cleaned data to another csv file:

```{r}
# Write to CSV file
write.csv(movies_clean, "movies_clean.csv", row.names = FALSE)

cat("Cleaned data saved as 'movies_clean.csv'\n")
cat("File location:", getwd(), "\n")

```

```{r check-new-file}
movies_clean_data <- read_csv('movies_clean.csv')
str(movies_clean_data)

```

**Data Cleaning of Movie_History dataset of Netflix - (Mehreen Gillani)**

### Step 2. Read watch_history dataset and store it in a Dataframe

```{r read_csv}
watch_history_url = 'https://raw.githubusercontent.com/mehreengillani/DATA607_project3/refs/heads/main/watch_history.csv'
watch_history = read.csv(watch_history_url)
colnames(watch_history)

#data structure
str(watch_history)
```

### Step 3: Data Cleaning

####  3.1 Handling missing values
```{r identify missing vals}

# Identify missing values
miss_var_summary(watch_history) %>%
  gt() %>%
  gt_theme_guardian() %>%
  tab_header(title="Missing data in watch_history table")

#plot
gg_miss_var(watch_history)
``` 

Immediate Assessment

High missingness (>79%): user_rating I will delete this column
for watch_duration_minutes, progress_percentage missing variable percentage is 11.7 and 8.11 so we will replace missing values with multiple imputation 

```{r user_rating column delete}

# Keep all columns, only impute the specific missing ones
watch_history_clean <- watch_history

# Create a temporary dataset for imputation
temp_impute <- watch_history_clean %>%
  select(watch_duration_minutes, progress_percentage, device_type, action, quality)

# Perform imputation
imputed_temp <- mice(temp_impute, m = 1, method = "pmm", printFlag = FALSE)
imputed_temp_complete <- complete(imputed_temp)

# Replace only the imputed columns in the main dataset
watch_history_clean$watch_duration_minutes <- imputed_temp_complete$watch_duration_minutes
watch_history_clean$progress_percentage <- imputed_temp_complete$progress_percentage

#check for missing data
miss_var_summary(watch_history_clean)
```

Check density plot after imputation to see if our distribution is still the same 
```{r}
# Density plots of imdb_ratings before and after imputation
par(mfrow = c(1, 2))
plot(density(watch_history$watch_duration_minutes, na.rm = TRUE), main = "watch_duration_minutes Density \n before multiple imputation")
plot(density(watch_history_clean$watch_duration_minutes, na.rm = TRUE), main = "watch_duration_minutes \n Density after imputation")

plot(density(watch_history$progress_percentage, na.rm = TRUE), main = "progress_percentage Density \n before multiple imputation")
plot(density(watch_history_clean$progress_percentage, na.rm = TRUE), main = "progress_percentage \n Density after imputation")

par(mfrow = c(1, 1))
```

Distribution is almost the same after imputation.

Lets Calculate before/after statistics

```{r}

# Calculate before/after statistics of watch_duration_minutes
before <- watch_history$watch_duration_minutes[!is.na(watch_history$watch_duration_minutes)]
after <- watch_history_clean$watch_duration_minutes

# Basic statistics comparison
stats_before <- c(
  mean = mean(before),
  sd = sd(before),
  median = median(before),
  min = min(before),
  max = max(before),
  n = length(before)
)

stats_after <- c(
  mean = mean(after),
  sd = sd(after),
  median = median(after),
  min = min(after),
  max = max(after),
  n = length(after)
)

comparison <- data.frame(Before = stats_before, After = stats_after, Change = stats_after - stats_before, percentage = ((stats_after - stats_before)/stats_before) * 100)
print(round(comparison, 3))
```

Watch Duration Minutes - Imputation Impact Summary

✅ Minimal Impact - Excellent Results

Mean: -0.471 minutes (-0.72%)

Negligible decrease in average watch time
Standard Deviation: -1.551 minutes (-2.27%)

Slight reduction in variability, actually beneficial
Median: +0.100 minutes (+0.20%)

Virtually unchanged central tendency
Range: Perfectly preserved

Min (0.2 min) and Max (799.3 min) unchanged
No artificial constraints introduced
Sample Size: +12,332 records (+13.3%)

Successfully recovered missing data points

```{r}

# Calculate before/after statistics of progress_percentage
before <- watch_history$progress_percentage[!is.na(watch_history$progress_percentage)]
after <- watch_history_clean$progress_percentage

# Basic statistics comparison
stats_before <- c(
  mean = mean(before),
  sd = sd(before),
  median = median(before),
  min = min(before),
  max = max(before),
  n = length(before)
)

stats_after <- c(
  mean = mean(after),
  sd = sd(after),
  median = median(after),
  min = min(after),
  max = max(after),
  n = length(after)
)

comparison <- data.frame(Before = stats_before, After = stats_after, Change = stats_after - stats_before, percentage = ((stats_after - stats_before)/stats_before) * 100)
print(round(comparison, 3))
```

Progress Percentage - Imputation Impact Summary

✅ Exceptional Preservation - Nearly Perfect

Mean: -0.121% (-0.24%)

Virtually identical average completion rate
Standard Deviation: -0.081% (-0.28%)

Minimal change in variability
Median: -0.100% (-0.20%)

Negligible shift in central tendency
Range: Perfectly preserved

Min (0%) and Max (100%) unchanged
Full scale integrity maintained
Sample Size: +8,514 records (+8.82%)

Significant recovery of missing progress data


```{r}
#check categorical values and see their unique values
unique(watch_history_clean$action)
unique(watch_history_clean$quality)
```
Replace NA with unknown 
```{r}
watch_history_clean1 <- watch_history_clean %>%
  mutate(
    # Action: convert NA to "unknown" ← THIS LINE HANDLES ACTION NA
    action = as.factor(ifelse(is.na(action), "unknown", as.character(action))),
    
    # Quality: preserve ordering, add "unknown" as lowest level
    quality = case_when(
      is.na(quality) ~ "unknown",
      TRUE ~ as.character(quality)
    ),
    quality = factor(quality, levels = c("unknown", "SD", "HD", "4K", "HDR"), ordered = TRUE)
  )
unique(watch_history_clean$action)
unique(watch_history_clean$quality)
```

#### Step 3.2: Removing Duplicates

```{r check_duplicates}
# Identify total duplicates
sum(duplicated(watch_history_clean1))
duplicated_percent <- (sum(duplicated(watch_history_clean1)) / nrow(watch_history_clean1) * 100)
print(duplicated_percent)

```

The percentage of duplicates is 4.07 so we will delete duplicated rows

```{r removeduplicates}

# Remove ALL duplicate rows (keeping first occurrence)
watch_history_unique <- distinct(watch_history_clean1)

# Verify removal
print(paste("Before:", nrow(watch_history_clean1), "rows"))
print(paste("After:", nrow(watch_history_unique), "rows"))
print(paste("Removed:", nrow(watch_history_clean1) - nrow(watch_history_unique), "duplicate rows"))
```

####  Step: 3.3 Data Type Conversion

```{r check data types}
# Basic structure
str(watch_history_unique)
```
```{r}
unique(watch_history_unique$quality)
```

```{r data type_conversion}


watch_history_final <- watch_history_unique %>%
  mutate(
    # Date conversion
    watch_date = as.Date(watch_date),
    
    # Logical conversion
    is_download = as.logical(is_download),
    
    # Ordered factors with logical hierarchies
    device_type = factor(device_type, 
                        levels = c("Mobile", "Tablet", "Laptop", "Desktop", "Smart TV"),
                        ordered = TRUE),
    
    # FIXED: Removed trailing comma after "completed"
    action = factor(action,
                   levels = c("unknown", "stopped", "paused", "started", "completed"),
                   ordered = TRUE),
    
    quality = factor(quality,
                    levels = c("unknown", "SD", "HD", "4K", "HDR"),
                    ordered = TRUE),
    
    # Regular factors
    location_country = as.factor(location_country)
  )
  
```

We have converted watch_date to date format, is_download logical and
categorical values to factors:
device_type, action, genre_secondary, quality, location_country

```{r}
#check new structure 
str(watch_history_final)
```

####  Step 3.4. Handling Outliers

```{r fine-numeric-columns}
# Find numeric columns
numeric_cols <- names(watch_history_final)[sapply(watch_history_final, is.numeric)]

cat("Numeric columns:\n")
print(numeric_cols)

```


BoxPlot before Outlier handling 

```{r beforeoutlierdetection}

boxplot(watch_history_final$watch_duration_minutes, main = "watch_duration_minutes Outliers", col = "lightblue")
boxplot(watch_history_final$progress_percentage, main = "progress_percentage Outliers", col = "lightgreen")

```
From these plots we can see there are no outliers for progress_percentage. But there are in watch_duration_minutes

Create new column duration_category in dataset

```{r}
watch_history_final <- watch_history_final %>%
  mutate(
    duration_category = factor(
      case_when(
        watch_duration_minutes == 0 ~ "0 minutes",
        watch_duration_minutes <= 5 ~ "1-5 minutes",
        watch_duration_minutes <= 30 ~ "6-30 minutes",
        watch_duration_minutes <= 60 ~ "31-60 minutes",
        watch_duration_minutes <= 120 ~ "1-2 hours",
        watch_duration_minutes <= 240 ~ "2-4 hours", 
        watch_duration_minutes <= 480 ~ "4-8 hours",
        TRUE ~ "Other"
      ),
      levels = c("0 minutes", "1-5 minutes", "6-30 minutes", "31-60 minutes", 
                 "1-2 hours", "2-4 hours", "4-8 hours", "Other"),
      ordered = TRUE
    )
  )

```

```{r}
# Calculate outlier bounds using IQR method
duration_data <- watch_history_final$watch_duration_minutes

Q1 <- quantile(duration_data, 0.25)
Q3 <- quantile(duration_data, 0.75)
IQR <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

cat("=== OUTLIER DETECTION ===\n")
cat("Q1 (25th percentile):", Q1, "minutes\n")
cat("Q3 (75th percentile):", Q3, "minutes\n") 
cat("IQR:", IQR, "minutes\n")
cat("Lower bound:", lower_bound, "minutes\n")
cat("Upper bound:", upper_bound, "minutes\n")

outliers <- sum(duration_data > upper_bound)
cat("Outliers detected:", outliers, "(", round(outliers/length(duration_data)*100, 2), "%)\n")


```
Analysis of Results

Key Issues Identified:
❌ Lower bound is negative (-50.5 minutes) - impossible for watch duration
❌ Upper bound seems too low (161.5 minutes = 2.7 hours)
✅ Only 3.83% outliers - manageable amount

Set lower value to 0 and upper value to 8 hours

```{r}

# Set the bounds
LOWER_BOUND <- 0      # Minimum 0 minutes
UPPER_BOUND <- 8 * 60 # Maximum 8 hours (480 minutes)

# Apply the capping
watch_history_final <- watch_history_final %>%
  mutate(
    watch_duration_minutes = case_when(
      watch_duration_minutes < LOWER_BOUND ~ LOWER_BOUND,
      watch_duration_minutes > UPPER_BOUND ~ UPPER_BOUND,
      TRUE ~ watch_duration_minutes
    )
  )

# Calculate statistics
original_data <- watch_history_final$watch_duration_minutes  # Before any changes if you saved it
capped_below <- sum(watch_history_final$watch_duration_minutes < LOWER_BOUND, na.rm = TRUE)
capped_above <- sum(watch_history_final$watch_duration_minutes > UPPER_BOUND, na.rm = TRUE)

cat("=== BOUNDS CAPPING RESULTS ===\n")
cat("Lower bound:", LOWER_BOUND, "minutes\n")
cat("Upper bound:", UPPER_BOUND, "minutes (", UPPER_BOUND/60, "hours)\n")
cat("Values capped below:", capped_below, "\n")
cat("Values capped above:", capped_above, "\n")
cat("New range: [", min(watch_history_final$watch_duration_minutes), ", ", 
    max(watch_history_final$watch_duration_minutes), "] minutes\n", sep = "")
```



BoxPlot after outlier handling with cap method

```{r after outlier cap}

boxplot(watch_history_final$watch_duration_minutes, main = "watch_duration_minutes after applying cap", col = "lightblue")

# Reset plot layout
par(mfrow = c(1, 1))
```

```{r}

# Simple bar plot
ggplot(watch_history_final, aes(x = duration_category)) +
  geom_bar(fill = "steelblue", alpha = 0.7) +
  labs(title = "Watch Duration Categories",
       x = "Duration Category", 
       y = "Number of Sessions") +
  theme_minimal()
```


####  3.5 String Cleaning


```{r findtextCols_string_cleaning}
# Find all text columns
text_cols <- names(watch_history_final)[sapply(watch_history_final, function(x) is.character(x) | is.factor(x))]

cat("Text columns found:\n")
print(text_cols)
```

Text cleaning: convert to lower case, convert empty string, remove whitespace

```{r textcleaning_p2}
# Perform comprehensive text cleaning
watch_history_cleaned <- watch_history_final %>%
  mutate(across(all_of(text_cols), ~ {
    if(is.character(.)) {
      # Remove leading/trailing whitespace
      cleaned <- trimws(.)
      # Convert to lowercase for standardization
      cleaned <- tolower(cleaned)
      # Remove multiple spaces
      cleaned <- gsub("\\s+", " ", cleaned)
      # Convert empty strings to NA
      cleaned <- ifelse(cleaned == "", NA, cleaned)
      return(cleaned)
    } else {
      return(.)
    }
  }))
```


```{r}
colnames(watch_history_cleaned)

unique(watch_history_cleaned$quality)
```

####  3.6 Data Validation

```{r Validatevariables}

# Comprehensive data validation function
validate_cleaned_data <- function(df) {
  cat("=== COMPREHENSIVE DATA VALIDATION REPORT ===\n\n")
  
  # 1. Basic Structure Validation
  cat("1. BASIC STRUCTURE VALIDATION\n")
  cat("   Total rows:", nrow(df), "\n")
  cat("   Total columns:", ncol(df), "\n")
  cat("   Complete cases:", sum(complete.cases(df)), "(", 
      round(mean(complete.cases(df)) * 100, 1), "%)\n")
  
  duplicates <- sum(duplicated(df))
  cat("   Duplicate rows:", duplicates, "\n")
  
  # 2. Missing Values Check
  cat("\n2. MISSING VALUES VALIDATION\n")
  missing_summary <- sapply(df, function(x) sum(is.na(x)))
  missing_df <- data.frame(
    Variable = names(missing_summary),
    Missing_Count = missing_summary,
    Missing_Percent = round(missing_summary / nrow(df) * 100, 2)
  ) %>% arrange(desc(Missing_Percent))
  
  print(missing_df)
  
  # 3. Data Type Validation
cat("\n3. DATA TYPE VALIDATION\n")

# Define expected validation functions for each column
expected_validations <- list(
  session_id = function(x) is.character(x),
  user_id = function(x) is.character(x),
  movie_id = function(x) is.character(x),
  watch_date = function(x) inherits(x, "Date"),
  device_type = function(x) is.character(x) || is.factor(x),
  watch_duration_minutes = function(x) is.numeric(x),
  progress_percentage = function(x) is.numeric(x),
  action = function(x) is.character(x) || is.factor(x),
  quality = function(x) is.character(x) || is.factor(x),
  location_country = function(x) is.character(x) || is.factor(x),
  is_download = function(x) is.logical(x),
  user_rating = function(x) is.numeric(x),
  duration_category = function(x) is.character(x) || is.factor(x)
)

# Build validation table
type_validation <- data.frame(
  Column = names(expected_validations),
  Data_Type = sapply(names(expected_validations), function(col) {
    if (col %in% names(df)) paste(class(df[[col]]), collapse = ", ") else NA
  }),
  Valid = sapply(names(expected_validations), function(col) {
    if (!col %in% names(df)) return(FALSE)
    expected_validations[[col]](df[[col]])
  }),
  stringsAsFactors = FALSE
)

# Print full validation table
print(type_validation)

# Optional: Highlight invalid columns
invalid_columns <- type_validation[type_validation$Valid == FALSE, ]
if (nrow(invalid_columns) > 0) {
  cat("\n⚠️ Columns with invalid types:\n")
  print(invalid_columns)
} else {
  cat("\n✅ All columns passed type validation.\n")
}

  
  # 4. Value Range Validation
  cat("\n4. VALUE RANGE VALIDATION\n")
  
  # watch_duration_minutes
  duration_stats <- c(
    Min = min(df$watch_duration_minutes),
    Max = max(df$watch_duration_minutes),
    Mean = mean(df$watch_duration_minutes),
    Negative_Values = sum(df$watch_duration_minutes < 0)
  )
  cat("   watch_duration_minutes:\n")
  cat("     Range: [", duration_stats["Min"], ", ", duration_stats["Max"], "]\n", sep = "")
  cat("     Negative values:", duration_stats["Negative_Values"], "\n")
  
  # progress_percentage
  progress_stats <- c(
    Min = min(df$progress_percentage),
    Max = max(df$progress_percentage),
    Out_of_Bounds = sum(df$progress_percentage < 0 | df$progress_percentage > 100)
  )
  cat("   progress_percentage:\n")
  cat("     Range: [", progress_stats["Min"], ", ", progress_stats["Max"], "]\n", sep = "")
  cat("     Values outside 0-100%:", progress_stats["Out_of_Bounds"], "\n")
  
  # 5. Categorical Value Validation
  cat("\n5. CATEGORICAL VALUE VALIDATION\n")
  
  categorical_vars <- c("device_type", "action", "quality", "location_country", "duration_category")
  for(var in categorical_vars) {
    unique_vals <- unique(df[[var]])
    cat("   ", var, ": ", length(unique_vals), " unique values\n", sep = "")
    
    # Check for empty strings
    if(is.character(df[[var]])) {
      empty_count <- sum(df[[var]] == "" | is.na(df[[var]]), na.rm = TRUE)
      if(empty_count > 0) cat("     Empty/NA values:", empty_count, "\n")
    }
  }
  
  # 6. Logical Variable Check
  cat("\n6. LOGICAL VARIABLE VALIDATION\n")
  if("is_download" %in% names(df)) {
    download_stats <- table(df$is_download, useNA = "always")
    cat("   is_download values:\n")
    print(download_stats)
  }
  
  # 7. Date Validation
  cat("\n7. DATE VALIDATION\n")
  if("watch_date" %in% names(df)) {
    date_range <- range(df$watch_date, na.rm = TRUE)
    cat("   watch_date range:", as.character(date_range[1]), "to", as.character(date_range[2]), "\n")
    
    future_dates <- sum(df$watch_date > Sys.Date(), na.rm = TRUE)
    cat("   Future dates:", future_dates, "\n")
  }
  
  # 8. Business Logic Validation
  cat("\n8. BUSINESS LOGIC VALIDATION\n")
  logic_checks <- data.frame(
    Check = character(),
    Violations = integer(),
    stringsAsFactors = FALSE
  )
  
  # Progress should not exceed 100% for non-completed actions
  if("progress_percentage" %in% names(df) && "action" %in% names(df)) {
    high_progress_not_completed <- sum(df$progress_percentage > 90 & df$action != "completed", na.rm = TRUE)
    logic_checks <- rbind(logic_checks, data.frame(
      Check = "High progress but not completed",
      Violations = high_progress_not_completed
    ))
  }
  
  # Watch duration should be positive
  negative_duration <- sum(df$watch_duration_minutes < 0, na.rm = TRUE)
  logic_checks <- rbind(logic_checks, data.frame(
    Check = "Negative watch duration",
    Violations = negative_duration
  ))
  
  print(logic_checks)
  
  # 9. Calculate Overall Quality Score
  cat("\n9. OVERALL DATA QUALITY SCORE\n")
  
  quality_metrics <- c(
    completeness = mean(complete.cases(df)),
    uniqueness = 1 - (duplicates / nrow(df)),
    valid_ranges = 1 - ((duration_stats["Negative_Values"] + progress_stats["Out_of_Bounds"]) / (nrow(df) * 2)),
    business_rules = 1 - (sum(logic_checks$Violations) / (nrow(df) * nrow(logic_checks)))
  )
  
  # Handle any NaN values
  quality_metrics <- sapply(quality_metrics, function(x) ifelse(is.nan(x), 1, x))
  
  quality_score <- mean(quality_metrics) * 100
  
  # Display quality metrics
  for(i in seq_along(quality_metrics)) {
    cat("   ", names(quality_metrics)[i], ": ", round(quality_metrics[i] * 100, 1), "%\n", sep = "")
  }
  
  cat("\n   OVERALL DATA QUALITY SCORE: ", round(quality_score, 1), "%\n", sep = "")
  
  # Quality rating
  if(quality_score >= 95) {
    rating <- "EXCELLENT"
    emoji <- "🎯"
  } else if(quality_score >= 85) {
    rating <- "VERY GOOD" 
    emoji <- "✅"
  } else if(quality_score >= 75) {
    rating <- "GOOD"
    emoji <- "✓"
  } else if(quality_score >= 65) {
    rating <- "ACCEPTABLE"
    emoji <- "⚠️"
  } else {
    rating <- "NEEDS IMPROVEMENT"
    emoji <- "❌"
  }
  
  cat("   ", emoji, " DATA QUALITY: ", rating, "\n\n", sep = "")
  
  # 10. Quick Visual Checks
  cat("10. VISUAL DISTRIBUTION CHECKS\n")
  
  # Plot duration categories
  p1 <- ggplot(df, aes(x = duration_category)) +
    geom_bar(fill = "steelblue", alpha = 0.7) +
    labs(title = "Duration Categories", x = "", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Plot watch duration distribution
  p2 <- ggplot(df, aes(x = watch_duration_minutes)) +
    geom_histogram(fill = "coral", alpha = 0.7, bins = 30) +
    labs(title = "Watch Duration Distribution", x = "Minutes", y = "Count") +
    theme_minimal()
  
  print(p1)
  print(p2)
  
  return(list(
    quality_score = quality_score,
    rating = rating,
    missing_summary = missing_df,
    logic_violations = logic_checks
  ))
}

# Run the validation
validation_results <- validate_cleaned_data(watch_history_cleaned)

```



This validation step is crucial before any analysis to ensure your results are reliable and meaningful!

Write cleaned data to another csv file:

```{r}
# Write to CSV file
write.csv(watch_history_cleaned, "watch_history_clean.csv", row.names = FALSE)

cat("Cleaned data saved as 'watch_history_clean.csv'\n")
cat("File location:", getwd(), "\n")

```

```{r check_new_file}
watch_history_cleaned_data <- read_csv('watch_history_clean.csv')
#str(watch_history_cleaned_data)

```
**Data Cleaning of Recommendation_Log dataset of Netflix - (Mehreen Gillani)**

### Step 2. Read recommendation_logs dataset and store it in a Dataframe

```{r readcsv}
recommendation_logs_url = 'https://raw.githubusercontent.com/mehreengillani/DATA607_project3/refs/heads/main/recommendation_logs.csv'
recommendation_logs = read.csv(recommendation_logs_url)
#column names
colnames(recommendation_logs)
#dimension 
dim(recommendation_logs)
#data structure
str(recommendation_logs)
```

### Step 3: Data Cleaning

####  3.1 Handling missing values
```{r ident_missing_vals}

# Identify missing values
miss_var_summary(recommendation_logs) %>%
  gt() %>%
  gt_theme_guardian() %>%
  tab_header(title="Missing data in recommendation_logs table")

#plot
gg_miss_var(recommendation_logs)
``` 

Immediate Assessment

Onlyone columns has missing values, recommendation_score	5216	10%

```{r recommendation_score missing values}


# Use other variables to predict missing scores
temp_df <- recommendation_logs %>%
  select(recommendation_score, was_clicked, position_in_list, device_type)

imputed_temp <- mice(temp_df, m = 1, method = "pmm", printFlag = FALSE)
imputed_temp_complete <- complete(imputed_temp)

recommendation_clean <- recommendation_logs
recommendation_clean$recommendation_score <- imputed_temp_complete$recommendation_score

# Final check
miss_var_summary(recommendation_clean)
```

Check density plot after imputation to see if our distribution is still the same 
```{r}
# Density plots of imdb_ratings before and after imputation
par(mfrow = c(1, 2))
plot(density(recommendation_logs$recommendation_score, na.rm = TRUE), main = "recommendation_score Density \n before multiple imputation")
plot(density(recommendation_clean$recommendation_score, na.rm = TRUE), main = "recommendation_score Density \n after imputation")

par(mfrow = c(1, 1))
```

Distribution is almost the same after imputation. It has minor changes in the graph

Lets Calculate before/after statistics

```{r before after missing value statstics}

# Calculate before/after statistics of \n 
before <- recommendation_logs$recommendation_score[!is.na(recommendation_logs$recommendation_score)]
after <- recommendation_clean$recommendation_score

# Basic statistics comparison
stats_before <- c(
  mean = mean(before),
  sd = sd(before),
  median = median(before),
  min = min(before),
  max = max(before),
  n = length(before)
)

stats_after <- c(
  mean = mean(after),
  sd = sd(after),
  median = median(after),
  min = min(after),
  max = max(after),
  n = length(after)
)

comparison <- data.frame(Before = stats_before, After = stats_after, Change = stats_after - stats_before, percentage = ((stats_after - stats_before)/stats_before) * 100)
print(round(comparison, 3))
```

recommendation_score - Imputation Impact Summary

✅ Minimal Impact - Excellent Results

Mean: change 0.005 (0.899%)

Negligible decrease in recommendation_score
Standard Deviation: -0.004 minutes (-1.684%)

No change in 
Median, min, max values 

No artificial constraints introduced
Sample Size: +5216 records (+11.1%)

Successfully recovered missing data points


```{r}
#check categorical values and see their unique values
cat("recommendation_type:\n")
print(unique(recommendation_clean$recommendation_type))

cat("\ndevice_type:\n") 
print(unique(recommendation_clean$device_type))

cat("\nwas_clicked:\n")
print(unique(recommendation_clean$was_clicked))
```

#### Step 3.2: Removing Duplicates

```{r ck duplicates}
# Identify total duplicates
sum(duplicated(recommendation_clean))
duplicated_percent <- (sum(duplicated(recommendation_clean)) / nrow(recommendation_clean) * 100)
print(duplicated_percent)

```

The percentage of duplicates is 3.525 so we will delete duplicated rows

```{r remove dupe}

# Remove ALL duplicate rows (keeping first occurrence)
recommendation_logs_unique <- distinct(recommendation_clean)

# Verify removal
print(paste("Before:", nrow(recommendation_clean), "rows"))
print(paste("After:", nrow(recommendation_logs_unique), "rows"))
print(paste("Removed:", nrow(recommendation_clean) - nrow(recommendation_logs_unique), "duplicate rows"))
```

####  Step: 3.3 Data Type Conversion

```{r check_data types}
# Basic structure
str(recommendation_logs_unique)
```

```{r data-type-conversion}


# If you want better ordering for analysis
recommendation_logs_final <- recommendation_logs_unique %>%
  mutate(
    # Date conversion
    recommendation_date = as.Date(recommendation_date),
    
    # Logical conversion
    was_clicked = as.logical(was_clicked),
    
    # Ordered factors with meaningful hierarchies
    recommendation_type = factor(recommendation_type,
                               levels = c("personalized", "genre_based", "trending", "new_releases", "similar_users"),
                               ordered = TRUE),
    
    device_type = factor(device_type,
                        levels = c("Mobile", "Tablet", "Laptop", "Desktop", "Smart TV"),
                        ordered = TRUE),
    
    time_of_day = factor(time_of_day,
                        levels = c("morning", "afternoon", "evening", "night"),
                        ordered = TRUE),
    
    algorithm_version = factor(algorithm_version)  # Regular factor
  )
  
```

We have converted recommendation_date to date format, was_clicked logical and
categorical values to factors:
recommendation_type, device_type, time_of_day, algorithm_version

```{r}
#check new structure 
str(recommendation_logs_final)
```

####  Step 3.4. Handling Outliers

```{r finenumeric columns}
# Find numeric columns
numeric_cols <- names(recommendation_logs_final)[sapply(recommendation_logs_final, is.numeric)]

cat("Numeric columns:\n")
print(numeric_cols)

```


BoxPlot before Outlier handling 

```{r b4_outlier_detection}

boxplot(recommendation_logs_final$recommendation_score, main = "recommendation_score Outliers", col = "lightblue")
boxplot(recommendation_logs_final$position_in_list, main = "position_in_list Outliers", col = "lightgreen")

```
From these plots we can see there are no outliers for recommendation_score and for position_in_list.  


####  3.5 String Cleaning


```{r find textCols_string_cleaning}
# Find all text columns
text_cols <- names(recommendation_logs_final)[sapply(recommendation_logs_final, function(x) is.character(x) | is.factor(x))]

cat("Text columns found:\n")
print(text_cols)
```

Text cleaning: convert to lower case, convert empty string, remove whitespace

```{r textcleaningp2}
# Perform comprehensive text cleaning
recommendation_logs_clean <- recommendation_logs_final %>%
  mutate(across(all_of(text_cols), ~ {
    if(is.character(.)) {
      # Remove leading/trailing whitespace
      cleaned <- trimws(.)
      # Convert to lowercase for standardization
      cleaned <- tolower(cleaned)
      # Remove multiple spaces
      cleaned <- gsub("\\s+", " ", cleaned)
      # Convert empty strings to NA
      cleaned <- ifelse(cleaned == "", NA, cleaned)
      return(cleaned)
    } else {
      return(.)
    }
  }))
```


```{r}
colnames(recommendation_logs_clean)

```

####  3.6 Data Validation

```{r Validate_variables}
validate_cleaned_data <- function(df) {
  cat("=== COMPREHENSIVE DATA VALIDATION REPORT ===\n\n")
  
  # 1. Basic Structure Validation
  cat("1. BASIC STRUCTURE VALIDATION\n")
  cat("   Total rows:", nrow(df), "\n")
  cat("   Total columns:", ncol(df), "\n")
  cat("   Complete cases:", sum(complete.cases(df)), "(", 
      round(mean(complete.cases(df)) * 100, 1), "%)\n")
  
  duplicates <- sum(duplicated(df))
  cat("   Duplicate rows:", duplicates, "\n")
  
  # 2. Missing Values Check
  cat("\n2. MISSING VALUES VALIDATION\n")
  missing_summary <- sapply(df, function(x) sum(is.na(x)))
  missing_df <- data.frame(
    Variable = names(missing_summary),
    Missing_Count = missing_summary,
    Missing_Percent = round(missing_summary / nrow(df) * 100, 2)
  ) %>% arrange(desc(Missing_Percent))
  
  print(missing_df)
  
  # 3. Data Type Validation (FIXED)
  cat("\n3. DATA TYPE VALIDATION\n")
  
  # Get data types safely
  data_types <- sapply(df, function(x) paste(class(x), collapse = ", "))
  type_validation <- data.frame(
    Column = names(data_types),
    Data_Type = as.character(data_types),
    stringsAsFactors = FALSE
  )
  print(type_validation)
  
  # 4. Value Range Validation
  cat("\n4. VALUE RANGE VALIDATION\n")
  
  # recommendation_score
  if("recommendation_score" %in% names(df)) {
    score_stats <- c(
      Min = min(df$recommendation_score, na.rm = TRUE),
      Max = max(df$recommendation_score, na.rm = TRUE),
      Mean = mean(df$recommendation_score, na.rm = TRUE),
      Out_of_Bounds = sum(df$recommendation_score < 0 | df$recommendation_score > 1, na.rm = TRUE)
    )
    cat("   recommendation_score:\n")
    cat("     Range: [", score_stats["Min"], ", ", score_stats["Max"], "]\n", sep = "")
    cat("     Values outside 0-1:", score_stats["Out_of_Bounds"], "\n")
  }
  
  # position_in_list
  if("position_in_list" %in% names(df)) {
    position_stats <- c(
      Min = min(df$position_in_list, na.rm = TRUE),
      Max = max(df$position_in_list, na.rm = TRUE),
      Negative_Values = sum(df$position_in_list < 0, na.rm = TRUE)
    )
    cat("   position_in_list:\n")
    cat("     Range: [", position_stats["Min"], ", ", position_stats["Max"], "]\n", sep = "")
    cat("     Negative values:", position_stats["Negative_Values"], "\n")
  }
  
  # 5. Categorical Value Validation
  cat("\n5. CATEGORICAL VALUE VALIDATION\n")
  
  categorical_vars <- names(df)[sapply(df, function(x) is.factor(x) | is.character(x))]
  # Remove ID columns from categorical analysis
  categorical_vars <- categorical_vars[!grepl("_id$", categorical_vars)]
  
  for(var in categorical_vars) {
    unique_vals <- unique(df[[var]])
    cat("   ", var, ": ", length(unique_vals), " unique values\n", sep = "")
    
    # Show top 5 values if not too many
    if(length(unique_vals) <= 10) {
      cat("     Values: ", paste(unique_vals, collapse = ", "), "\n", sep = "")
    }
    
    # Check for empty strings
    if(is.character(df[[var]])) {
      empty_count <- sum(df[[var]] == "" | is.na(df[[var]]), na.rm = TRUE)
      if(empty_count > 0) cat("     Empty/NA values:", empty_count, "\n")
    }
  }
  
  # 6. Logical Variable Check
  cat("\n6. LOGICAL VARIABLE VALIDATION\n")
  if("was_clicked" %in% names(df)) {
    click_stats <- table(df$was_clicked, useNA = "always")
    cat("   was_clicked values:\n")
    print(click_stats)
  }
  
  # 7. Date Validation
  cat("\n7. DATE VALIDATION\n")
  if("recommendation_date" %in% names(df)) {
    date_range <- range(df$recommendation_date, na.rm = TRUE)
    cat("   recommendation_date range:", as.character(date_range[1]), "to", as.character(date_range[2]), "\n")
    
    future_dates <- sum(df$recommendation_date > Sys.Date(), na.rm = TRUE)
    cat("   Future dates:", future_dates, "\n")
  }
  
  # 8. Business Logic Validation
  cat("\n8. BUSINESS LOGIC VALIDATION\n")
  logic_checks <- data.frame(
    Check = character(),
    Violations = integer(),
    stringsAsFactors = FALSE
  )
  
  # Recommendation score should be between 0-1
  if("recommendation_score" %in% names(df)) {
    invalid_scores <- sum(df$recommendation_score < 0 | df$recommendation_score > 1, na.rm = TRUE)
    logic_checks <- rbind(logic_checks, data.frame(
      Check = "Invalid recommendation scores (not 0-1)",
      Violations = invalid_scores
    ))
  }
  
  # Position should be positive
  if("position_in_list" %in% names(df)) {
    negative_positions <- sum(df$position_in_list < 0, na.rm = TRUE)
    logic_checks <- rbind(logic_checks, data.frame(
      Check = "Negative positions in list",
      Violations = negative_positions
    ))
  }
  
  print(logic_checks)
  
  # 9. Calculate Overall Quality Score
  cat("\n9. OVERALL DATA QUALITY SCORE\n")
  
  total_rows <- nrow(df)
  total_logic_checks <- max(nrow(logic_checks), 1)  # Avoid division by zero
  
  quality_metrics <- c(
    completeness = mean(complete.cases(df)),
    uniqueness = ifelse(total_rows > 0, 1 - (duplicates / total_rows), 1),
    valid_ranges = 1 - ((sum(df$recommendation_score < 0 | df$recommendation_score > 1, na.rm = TRUE) + 
                         sum(df$position_in_list < 0, na.rm = TRUE)) / (total_rows * 2)),
    business_rules = 1 - (sum(logic_checks$Violations) / (total_rows * total_logic_checks))
  )
  
  # Handle any NaN values
  quality_metrics <- sapply(quality_metrics, function(x) ifelse(is.nan(x), 1, x))
  
  quality_score <- mean(quality_metrics) * 100
  
  # Display quality metrics
  for(i in seq_along(quality_metrics)) {
    cat("   ", names(quality_metrics)[i], ": ", round(quality_metrics[i] * 100, 1), "%\n", sep = "")
  }
  
  cat("\n   OVERALL DATA QUALITY SCORE: ", round(quality_score, 1), "%\n", sep = "")
  
  # Quality rating
  if(quality_score >= 95) {
    rating <- "EXCELLENT"
    emoji <- "🎯"
  } else if(quality_score >= 85) {
    rating <- "VERY GOOD" 
    emoji <- "✅"
  } else if(quality_score >= 75) {
    rating <- "GOOD"
    emoji <- "✓"
  } else if(quality_score >= 65) {
    rating <- "ACCEPTABLE"
    emoji <- "⚠️"
  } else {
    rating <- "NEEDS IMPROVEMENT"
    emoji <- "❌"
  }
  
  cat("   ", emoji, " DATA QUALITY: ", rating, "\n\n", sep = "")
  
  return(list(
    quality_score = quality_score,
    rating = rating,
    missing_summary = missing_df,
    logic_violations = logic_checks
  ))
}

# Run the validation
validation_results <- validate_cleaned_data(recommendation_logs_clean)

```

This validation step is crucial before any analysis to ensure your results are reliable and meaningful!

Write cleaned data to another csv file:

```{r}
# Write to CSV file
write.csv(recommendation_logs_clean, "recommendation_logs_clean.csv", row.names = FALSE)

cat("Cleaned data saved as 'recommendation_logs_clean.csv'\n")
cat("File location:", getwd(), "\n")

```

```{r check new file}
recommendation_logs_clean_data <- read_csv('recommendation_logs_clean.csv')
#str(recommendation_logs_cleaned_data)

```

**Data visuals and analysis -- (Taha Malik)**

```{r setup, include=FALSE}
# Libraries (suppress startup noise)
suppressPackageStartupMessages({
  # optional libraries; code will fallback if missing
  if(!requireNamespace("vroom", quietly = TRUE)) library(readr)
  if(!requireNamespace("infer", quietly = TRUE)) NULL
})
set.seed(123)
DATA_DIR <- params$data_dir
OUT_DIR <- file.path(DATA_DIR, "output")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

Executive summary
-----------------

This report answers three primary evaluation questions:
1. How is user engagement distributed across the catalog (views, progress, duration)?  
2. Do recommended titles receive measurably higher engagement than non-recommended titles?  
3. How do recommendation algorithms and list position affect CTR?

Key findings (summary numbers will be filled automatically when you run the report):
- Recommended titles: mean progress = (auto-filled) vs non-recommended = (auto-filled). Statistical test reported in section 5.1.
- CTR shows strong position bias: position 1 CTR = (auto-filled) vs position 10 CTR = (auto-filled).
- Algorithm lift: logistic model reports adjusted odds ratios and AUC (see Recommendation logs section).

Actionable recommendations:
- Prioritize algorithm variants with significant lift in an A/B rollout, controlling for position.
- Use peak-hour serving and UI treatments informed by session-hour heatmap.
- Instrument financial fields more consistently (avoid treating unknown budgets as zero).

Notes on reproducibility
- This Rmd uses params$data_dir. To run locally, call rmarkdown::render("Project3.Rmd", params = list(data_dir = "path/to/dir")).
- Key artifacts are written to output/: top tables and publication plots.
- Session/package versions are saved at the end of the report.

---

# 0. Load cleaned datasets (robust and reproducible)

This section loads CSVs from params$data_dir using vroom for speed when available; otherwise uses readr::read_csv. If files are missing the report continues and sections that depend on them will be skipped with a clear message.

```{r load-cleaned, message=FALSE}
movies_file <- file.path(DATA_DIR, "movies_clean.csv")
users_file  <- file.path(DATA_DIR, "users_clean.csv")
watch_file  <- file.path(DATA_DIR, "watch_history_clean.csv")
rec_file    <- file.path(DATA_DIR, "recommendation_logs_clean.csv")

safe_read <- function(path){
  if(!file.exists(path)){
    message("File not found: ", path)
    return(tibble())
  }
  # prefer vroom for big files
  if(requireNamespace("vroom", quietly = TRUE)){
    vroom::vroom(path, progress = FALSE, show_col_types = FALSE)
  } else {
    readr::read_csv(path, show_col_types = FALSE)
  }
}

movies <- safe_read(movies_file) %>% as_tibble()
users  <- safe_read(users_file)  %>% as_tibble()
watch  <- safe_read(watch_file)  %>% as_tibble()
rec    <- safe_read(rec_file)    %>% as_tibble()

# Print shapes safely
cat("Shapes (rows x cols):\n")
cat("movies:", ifelse(nrow(movies)>0, paste(nrow(movies), "x", ncol(movies)), "missing"), "\n")
cat("users:",  ifelse(nrow(users)>0, paste(nrow(users), "x", ncol(users)), "missing"), "\n")
cat("watch_history:", ifelse(nrow(watch)>0, paste(nrow(watch), "x", ncol(watch)), "missing"), "\n")
cat("recommendation_logs:", ifelse(nrow(rec)>0, paste(nrow(rec), "x", ncol(rec)), "missing"), "\n")
```

Data dictionary & basic sanity checks
------------------------------------

This small table enumerates the columns present in each dataset and runs quick sanity checks (range checks, required ids). The generated CSV is saved to output/data_dictionary.csv.

```{r data-dictionary}
make_col_table <- function(df, name){
  if(nrow(df)==0) return(tibble(dataset = name, column = NA, type = NA, n_missing = NA))
  tibble(dataset = name,
         column = names(df),
         type = map_chr(df, ~ class(.x)[1]),
         n_missing = map_int(df, ~ sum(is.na(.x)))
         )
}
dict <- bind_rows(
  make_col_table(movies, "movies"),
  make_col_table(users, "users"),
  make_col_table(watch, "watch"),
  make_col_table(rec, "rec")
)
write_csv(dict, file.path(OUT_DIR, "data_dictionary.csv"))
kable(head(dict, 50), format = "html", table.attr = "style='width:100%'") %>% kable_styling()
```

Sanity checks (stop only if required id columns are missing)
```{r sanity-checks}
required_watch <- c("user_id","movie_id")
required_rec <- c("was_clicked")
if(nrow(watch)>0 && !all(required_watch %in% names(watch))){
  warning("watch_history missing required columns: ", paste(setdiff(required_watch, names(watch)), collapse = ", "))
}
if(nrow(rec)>0 && !all(required_rec %in% names(rec))){
  warning("recommendation_logs missing required columns: ", paste(setdiff(required_rec, names(rec)), collapse = ", "))
}
# Check ranges
if("progress_percentage" %in% names(watch)){
  if(any(watch$progress_percentage < 0 | watch$progress_percentage > 100, na.rm = TRUE)){
    warning("progress_percentage outside [0,100] detected")
  }
}
```

---

# 1. Movies — in-depth analysis

Purpose: understand catalog composition, ratings, runtime, and financials.

## 1.1 Overview & distributions (with accessibility and sample sizes)

```{r movies-overview, cache=TRUE}
if(nrow(movies) == 0){
  message("Skipping movies section: movies_clean.csv missing or empty")
} else {
  skim(movies)
  # imdb_rating distribution with sample size and accessible palette
  p1 <- ggplot(movies %>% filter(!is.na(imdb_rating)), aes(x = imdb_rating)) +
    geom_histogram(binwidth = 0.25, fill = viridis(1, option = "D"), color = "white", alpha = 0.95) +
    labs(title = "IMDb rating distribution", x = "IMDb rating", y = "Count",
         caption = paste0("n = ", sum(!is.na(movies$imdb_rating)))) +
    theme_minimal()
  # release year trend
  if("release_year" %in% names(movies)){
    p2 <- movies %>% filter(!is.na(release_year)) %>%
      count(release_year) %>%
      ggplot(aes(x = release_year, y = n)) +
      geom_col(fill = viridis(1, option = "C")) +
      labs(title = "Number of titles by release year", x = "Release year", y = "Count") +
      theme_minimal()
  } else {
    p2 <- ggplot() + ggtitle("release_year not available")
  }
  p1 + p2
}
```

Interpretation (example):
- Inspect skewness: if mean(imdb_rating) > median, likely right skew; comment with numbers will appear after knitting.

## 1.2 Runtime and rating relationships

```{r movies-runtime-rating, cache=TRUE}
if (nrow(movies) == 0) {
  message("Skipping runtime section")
} else if (!"duration_minutes" %in% names(movies)) {
  ggplot() + ggtitle("duration_minutes column not available")
} else {
  p_dur <- ggplot(movies, aes(x = duration_minutes)) +
    geom_histogram(bins = 60, fill = viridis(1, option = "B"), color = "white") +
    scale_x_continuous(labels = comma) +
    labs(
      title = "Distribution of runtime (minutes)",
      x = "Duration (minutes)",
      y = "Count",
      caption = paste0("n = ", sum(!is.na(movies$duration_minutes)))
    ) +
    theme_minimal()

  if ("rating" %in% names(movies)) {
    p_box <- ggplot(
      movies %>% filter(!is.na(rating) & !is.na(duration_minutes)),
      aes(x = rating, y = duration_minutes)
    ) +
      geom_boxplot(outlier.alpha = 0.3) +
      coord_flip() +
      labs(title = "Duration by rating (boxplot)", x = "", y = "Duration (minutes)") +
      theme_minimal()
  } else {
    p_box <- ggplot() + ggtitle("rating not available")
  }

  p_dur + p_box
}
```

## 1.3 Budget vs box-office revenue (financial analysis — avoid silently treating missing as zero)

We explicitly filter to rows where both budget and box office are non-missing. Treat zeros as valid only if documented (we print counts).

```{r movies-budget-revenue, cache=TRUE}
if(nrow(movies)==0){
  message("Skipping financials: movies missing")
} else if(all(c("production_budget","box_office_revenue") %in% names(movies))){
  # We will not coalesce to zero silently; only use complete pairs
  df_fin <- movies %>% filter(!is.na(production_budget) & !is.na(box_office_revenue)) %>%
    mutate(pb = production_budget, br = box_office_revenue,
           pb_log = log10(pb + 1), br_log = log10(br + 1),
           is_ng = ifelse(is_netflix_original, "Netflix Original", "Other"))
  cat("Rows with both budget and box-office:", nrow(df_fin), "\n")
  if(nrow(df_fin) > 5){
    p_scatter <- ggplot(df_fin, aes(x = pb_log, y = br_log, color = is_ng)) +
      geom_point(alpha = 0.6, size = 1) +
      geom_smooth(method = "loess", se = FALSE, color = "black") +
      labs(title = "log10(Production budget) vs log10(Box office revenue)",
           x = "log10(production_budget + 1)", y = "log10(box_office_revenue + 1)") +
      theme_minimal() + scale_color_viridis_d()
    p_box2 <- ggplot(df_fin %>% filter(!is.na(is_netflix_original)), aes(x = is_ng, y = br + 1)) +
      geom_boxplot() + scale_y_log10(labels = comma) +
      labs(title = "Box office revenue by origin (log scale)", x = "", y = "Box office (log)") + theme_minimal()
    p_scatter + p_box2
  } else {
    ggplot() + ggtitle("Insufficient financial data (please inspect output/data_dictionary.csv)")
  }
} else {
  ggplot() + ggtitle("Budget or revenue columns missing")
}
```

Interpretation & bias note:
- We explicitly mention if rows are filtered out due to missing financials. Do not treat unknown budgets as zero.

## 1.4 Genre analysis and top titles

```{r movies-genre-top, cache=TRUE}
if (nrow(movies) == 0) {
  message("Skipping genre section")
} else if ("genre_primary" %in% names(movies)) {
  top_genres <- movies %>% count(genre_primary, sort = TRUE) %>% head(15)

  p_gen <- ggplot(top_genres, aes(x = reorder(genre_primary, n), y = n)) +
    geom_col(fill = viridis(1, option = "D")) + coord_flip() +
    labs(
      title = "Top genres by title counts",
      x = "",
      y = "Count",
      caption = paste0("Total titles: ", nrow(movies))
    ) +
    theme_minimal()

  if ("imdb_rating" %in% names(movies)) {
    top_rated <- movies %>% filter(!is.na(imdb_rating)) %>% arrange(desc(imdb_rating)) %>% slice_head(n = 15)

    p_top <- ggplot(top_rated, aes(x = reorder(title, imdb_rating), y = imdb_rating)) +
      geom_col(fill = viridis(1, option = "C")) + coord_flip() +
      labs(title = "Top rated titles (by imdb_rating)", x = "", y = "IMDb rating") +
      theme_minimal()

    # display both panels side-by-side (patchwork)
    print(p_gen + p_top)
  } else {
    # only genre plot available
    print(p_gen)
  }
} else {
  print(ggplot() + ggtitle("genre_primary missing"))
}
```

Save top genre counts
```{r save_genre, eval = TRUE}
if("genre_primary" %in% names(movies)){
  write_csv(top_genres, file.path(OUT_DIR, "top_genres.csv"))
}
```

---

# 2. Users — demographics & linking to watch behavior

Goal: describe user base and compute activity metrics joined to watch history.

## 2.1 Users overview

```{r users-overview, cache=TRUE}
if (nrow(users) == 0) {
  message("Skipping users: users_clean.csv missing")
} else {
  skim(users)
  cats <- users %>%
    select(where(~ is.character(.x) || is.factor(.x))) %>%
    map(~ length(unique(.)))
  print(head(cats, 20))
}
```

## 2.2 Age and country distributions

```{r users-ages, cache=TRUE}
if(nrow(users)==0) ggplot() + ggtitle("users missing") else {
  p_age <- if("age" %in% names(users)) {
    ggplot(users %>% filter(!is.na(age)), aes(x = age)) +
      geom_histogram(binwidth = 2, fill = viridis(1, option = "B"), color = "white") +
      labs(title = "User age distribution", x = "Age", y = "Count",
           caption = paste0("n = ", sum(!is.na(users$age)))) + theme_minimal()
  } else ggplot() + ggtitle("age not available")

  p_country <- if("country" %in% names(users)) {
    users %>% count(country, sort = TRUE) %>% filter(!is.na(country)) %>% slice_head(n=20) %>%
      ggplot(aes(x = reorder(country, n), y = n)) + geom_col(fill = viridis(1, option = "D")) + coord_flip() +
      labs(title = "Top 20 user countries", x = "", y = "Users") + theme_minimal()
  } else ggplot() + ggtitle("country not available")

  p_age + p_country
}
```

## 2.3 Signup/cohort basics

```{r users-signup-cohort, cache=TRUE}
# Robust signup / proxy-from-first-watch chunk

safe_parse_datetime <- function(x) {
  # return POSIXct vector; try numeric epoch first, then common text formats
  x_chr <- as.character(x)
  n <- length(x_chr)
  out <- rep(NA_real_, n)

  # 1) try numeric epoch (seconds or milliseconds)
  num <- suppressWarnings(as.numeric(x_chr))
  is_num <- !is.na(num)
  if (any(is_num)) {
    # Heuristic: if large values, maybe milliseconds; convert if > 1e12
    num_vals <- num[is_num]
    num_vals[num_vals > 1e12] <- num_vals[num_vals > 1e12] / 1000
    out[is_num] <- num_vals
  }

  # 2) parse remaining with lubridate::parse_date_time using several common orders
  to_parse <- which(is.na(out) & !is.na(x_chr) & x_chr != "")
  if (length(to_parse) > 0) {
    parsed <- suppressWarnings(
      parse_date_time(x_chr[to_parse],
                      orders = c("ymd HMS", "ymd HM", "ymd", "mdy HMS", "mdy HM", "mdy",
                                 "dmy HMS", "dmy HM", "dmy", "Y-m-d H:M:S"),
                      tz = "UTC")
    )
    out[to_parse] <- as.numeric(parsed)
  }

  # 3) return POSIXct (as_datetime handles NA numeric)
  lubridate::as_datetime(out, tz = "UTC")
}

if (nrow(users) > 0 && "signup_date" %in% names(users)) {
  users <- users %>%
    mutate(signup_date = as_date(signup_date),
           signup_month = floor_date(signup_date, "month"))
  p_signup <- users %>%
    count(signup_month) %>%
    ggplot(aes(x = signup_month, y = n)) +
    geom_line(color = viridis(1, option = "C")) + geom_point() +
    labs(title = "Signups over time (monthly)", x = "Signup month", y = "Count",
         caption = paste0("Source: users$signup_date; total users = ", nrow(users))) +
    theme_minimal()
  print(p_signup)

} else if (nrow(users) > 0 && nrow(watch) > 0 && "watch_date" %in% names(watch) && "user_id" %in% names(watch)) {
  message("users$signup_date not found — deriving proxy signup_month from first watch_date per user (use with caution).")

  # create parsed datetime robustly
  watch <- watch %>%
    mutate(watch_date_parsed = safe_parse_datetime(watch_date))

  user_first_watch <- watch %>%
    filter(!is.na(watch_date_parsed) & !is.na(user_id)) %>%
    group_by(user_id) %>%
    summarise(first_watch = min(watch_date_parsed, na.rm = TRUE), .groups = "drop") %>%
    mutate(signup_month = floor_date(first_watch, "month"))

  # Save proxy table for transparency
  if (exists("OUT_DIR")) {
    write_csv(user_first_watch, file.path(OUT_DIR, "user_first_watch_proxy_signup.csv"))
  }

  p_proxy <- user_first_watch %>%
    count(signup_month) %>%
    ggplot(aes(x = signup_month, y = n)) +
    geom_line(color = viridis(1, option = "C")) + geom_point() +
    labs(title = "Proxy signups by first-watch month (proxy)",
         subtitle = "Proxy: first watch date per user used as signup approximation",
         x = "Proxy signup month", y = "Count",
         caption = paste0("n users with first-watch = ", nrow(user_first_watch))) +
    theme_minimal()
  print(p_proxy)

} else {
  message("Skipping signup/cohort analysis: no users$signup_date and no usable watch_date/user_id to derive a proxy.")
}
```

## 2.4 Join users to watch history: user activity metrics

```{r users-join-watch, cache=TRUE}
if(all(c("user_id") %in% names(users)) && all(c("user_id") %in% names(watch))){
  user_activity <- watch %>% group_by(user_id) %>%
    summarise(total_sessions = n(),
              mean_duration = mean(watch_duration_minutes, na.rm = TRUE),
              median_progress = median(progress_percentage, na.rm = TRUE), .groups = "drop")
  users_act <- users %>% left_join(user_activity, by = "user_id")
  if("age_group" %in% names(users_act)){
    p <- users_act %>% group_by(age_group) %>% summarise(mean_dur = mean(mean_duration, na.rm = TRUE), n = n()) %>%
      ggplot(aes(x = age_group, y = mean_dur)) + geom_col(fill = viridis(1, option = "B")) +
      labs(title = "Avg watch duration by age group", x = "Age group", y = "Avg duration (min)") + theme_minimal()
    print(p)
  }
  # Save top active users
  top_users <- users_act %>% arrange(desc(total_sessions)) %>% slice_head(n = 20)
  write_csv(top_users, file.path(OUT_DIR, "top_20_active_users.csv"))
  knitr::kable(top_users %>% select(user_id, total_sessions, mean_duration) %>% head(10)) %>% kable_styling()
} else {
  message("user_id missing in users or watch data; skip join analysis")
}
```

---

# 3. Watch history — session analysis & temporal patterns

Purpose: session-level engagement, device and time patterns.

## 3.1 Session-level distributions

```{r watch-session-dist, cache=TRUE}
if (nrow(watch) == 0) {
  message("Skipping watch session analysis")
} else {
  skim(watch)

  # Duration histogram (only if column exists)
  if ("watch_duration_minutes" %in% names(watch)) {
    p_dur <- ggplot(watch %>% filter(!is.na(watch_duration_minutes)),
                    aes(x = watch_duration_minutes)) +
      geom_histogram(bins = 80, fill = viridis(1, option = "B"), color = "white") +
      labs(
        title = "Watch duration distribution",
        x = "Duration (minutes)",
        y = "Sessions",
        caption = paste0("n = ", sum(!is.na(watch$watch_duration_minutes)))
      ) +
      theme_minimal()
  } else {
    p_dur <- ggplot() + ggtitle("watch_duration_minutes column not available")
  }

  # Progress histogram (only if column exists)
  if ("progress_percentage" %in% names(watch)) {
    p_prog <- ggplot(watch %>% filter(!is.na(progress_percentage)),
                     aes(x = progress_percentage)) +
      geom_histogram(bins = 40, fill = viridis(1, option = "D"), color = "white") +
      labs(
        title = "Progress percentage distribution",
        x = "Progress (%)",
        y = "Sessions",
        caption = paste0("n = ", sum(!is.na(watch$progress_percentage)))
      ) +
      theme_minimal()
  } else {
    p_prog <- ggplot() + ggtitle("progress_percentage column not available")
  }

  # Combine and print so it renders inside conditional
  print(p_dur + p_prog)
}
```

## 3.2 Device & action analysis

```{r watch-device-action, cache=TRUE}
if(nrow(watch)>0){
  if("device_type" %in% names(watch)){
    p_dev <- watch %>% count(device_type) %>% ggplot(aes(x = reorder(device_type, n), y = n)) +
      geom_col(fill = viridis(1, option = "C")) + coord_flip() + labs(title = "Sessions by device type", x = "", y = "Count") + theme_minimal()
    print(p_dev)
  }
  if(all(c("action","progress_percentage") %in% names(watch))){
    p_act <- watch %>% filter(!is.na(action)) %>% group_by(action) %>%
      summarise(mean_progress = mean(progress_percentage, na.rm = TRUE), n = n(), .groups = "drop") %>%
      ggplot(aes(x = reorder(action, mean_progress), y = mean_progress)) +
      geom_col(fill = viridis(1, option = "B")) + coord_flip() +
      labs(title = "Mean progress by action", x = "", y = "Mean progress (%)") + theme_minimal()
    print(p_act)
  }
}
```

## 3.3 Temporal patterns: hourly and daily

```{r watch-time-patterns, cache=TRUE}
# Robust watch-time plotting that handles Date (no time) vs datetime (with time)

safe_parse_datetime <- function(x) {
  # Try a few safe fallbacks; never throw an error
  x_chr <- as.character(x)
  out <- rep(NA_real_, length(x_chr))

  # 1) try base as.POSIXct (handles ISO-like strings)
  pos <- suppressWarnings(as.POSIXct(x_chr, tz = "UTC"))
  ok <- !is.na(pos)
  out[ok] <- as.numeric(pos[ok])

  # 2) parse remaining with lubridate::parse_date_time inside tryCatch
  rem <- which(is.na(out) & !is.na(x_chr) & x_chr != "")
  if (length(rem) > 0) {
    parsed <- tryCatch(
      parse_date_time(
        x_chr[rem],
        orders = c("ymd HMS","ymd HM","ymd","mdy HMS","mdy HM","mdy",
                   "dmy HMS","dmy HM","dmy","Y-m-d H:M:S","ISO8601"),
        tz = "UTC"
      ),
      error = function(e) rep(NA, length(rem))
    )
    out[rem] <- as.numeric(parsed)
  }

  # 3) final fallback: anytime::anytime if available
  rem2 <- which(is.na(out) & !is.na(x_chr) & x_chr != "")
  if (length(rem2) > 0 && requireNamespace("anytime", quietly = TRUE)) {
    parsed_any <- suppressWarnings(anytime::anytime(x_chr[rem2]))
    ok2 <- !is.na(parsed_any)
    out[rem2[ok2]] <- as.numeric(parsed_any[ok2])
  }

  as_datetime(out, tz = "UTC")
}

# Main chunk
if (nrow(watch) == 0) {
  message("Skipping watch-time-patterns: watch is empty")
} else if (!"watch_date" %in% names(watch)) {
  message("Skipping watch-time-patterns: watch_date column not present")
} else {
  wd <- watch$watch_date

  # CASE A: watch_date already a Date (no time-of-day)
  if (inherits(wd, "Date")) {
    message("watch_date is class Date (no time-of-day). Plotting daily and day-of-week counts.")
    watch <- watch %>% mutate(watch_day = as_date(watch_date))
    p_day <- watch %>% count(watch_day) %>%
      ggplot(aes(x = watch_day, y = n)) +
      geom_line(color = viridis::viridis(1)) + geom_point() +
      labs(title = "Sessions by day (date-level)", x = "Date", y = "Sessions",
           caption = paste0("n days = ", n_distinct(watch$watch_day))) +
      theme_minimal()
    p_dow <- watch %>% mutate(dow = wday(watch_day, label = TRUE)) %>% count(dow) %>%
      ggplot(aes(x = dow, y = n)) +
      geom_col(fill = viridis::viridis(1, option = "C")) +
      labs(title = "Sessions by day of week", x = "", y = "Sessions") + theme_minimal()
    print(p_day + p_dow)

  # CASE B: watch_date already has time component (POSIXct / POSIXlt)
  } else if (inherits(wd, "POSIXt")) {
    message("watch_date is POSIX datetime. Plotting hour-of-day and day-of-week.")
    watch <- watch %>% mutate(watch_datetime = as_datetime(watch_date),
                              hour = hour(watch_datetime),
                              dow = wday(watch_datetime, label = TRUE))
    hour_counts <- watch %>% filter(!is.na(hour)) %>% count(hour)
    p_hour <- ggplot(hour_counts, aes(x = hour, y = n)) +
      geom_col(fill = viridis::viridis(1)) + scale_x_continuous(breaks = 0:23) +
      labs(title = "Sessions by hour of day", x = "Hour (0-23)", y = "Sessions",
           caption = paste0("n parsed = ", sum(!is.na(watch$hour)))) + theme_minimal()
    p_dow <- watch %>% filter(!is.na(dow)) %>% count(dow) %>%
      ggplot(aes(x = factor(dow, levels = c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")), y = n)) +
      geom_col(fill = viridis::viridis(1, option = "C")) +
      labs(title = "Sessions by day of week", x = "", y = "Sessions") + theme_minimal()
    print(p_hour + p_dow)

  # CASE C: watch_date is character or other (attempt robust parse)
  } else {
    message("watch_date is not Date/POSIX; attempting robust parse. If parsed datetimes have no times, will fallback to date-level plots.")
    watch <- watch %>% mutate(watch_datetime = safe_parse_datetime(watch$watch_date))

    n_parsed <- sum(!is.na(watch$watch_datetime))
    message(sprintf("Parsed datetime count: %d / %d", n_parsed, nrow(watch)))

    # Save problematic raw samples for inspection
    bad_raw <- unique(as.character(watch$watch_date[is.na(watch$watch_datetime) & !is.na(watch$watch_date) & watch$watch_date != ""]))
    if (length(bad_raw) > 0 && exists("OUT_DIR")) {
      write_lines(bad_raw, file.path(OUT_DIR, "watch_date_unparseable_samples.txt"))
      message("Saved unparseable watch_date samples to output/watch_date_unparseable_samples.txt")
    }

    # If any parsed datetimes include non-midnight times, produce hourly plots
    if (n_parsed > 0 && any(lubridate::hour(watch$watch_datetime) != 0, na.rm = TRUE)) {
      watch <- watch %>% mutate(hour = hour(watch_datetime), dow = wday(watch_datetime, label = TRUE))
      hour_counts <- watch %>% filter(!is.na(hour)) %>% count(hour)
      p_hour <- ggplot(hour_counts, aes(x = hour, y = n)) + geom_col(fill = viridis::viridis(1)) + scale_x_continuous(breaks = 0:23) +
        labs(title = "Sessions by hour of day (parsed)", x = "Hour (0-23)", y = "Sessions") + theme_minimal()
      p_dow <- watch %>% filter(!is.na(dow)) %>% count(dow) %>%
        ggplot(aes(x = factor(dow, levels = c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")), y = n)) +
        geom_col(fill = viridis::viridis(1, option = "C")) +
        labs(title = "Sessions by day of week (parsed)", x = "", y = "Sessions") + theme_minimal()
      print(p_hour + p_dow)
    } else {
      # fallback: use date-only summaries
      watch <- watch %>% mutate(watch_day = as_date(watch_datetime))
      p_day <- watch %>% count(watch_day) %>%
        ggplot(aes(x = watch_day, y = n)) + geom_line(color = viridis::viridis(1)) + geom_point() +
        labs(title = "Sessions by day (date-level, parsed)", x = "Date", y = "Sessions") + theme_minimal()
      p_dow <- watch %>% mutate(dow = wday(watch_day, label = TRUE)) %>% count(dow) %>%
        ggplot(aes(x = dow, y = n)) + geom_col(fill = viridis::viridis(1, option = "C")) +
        labs(title = "Sessions by day of week (date-level)", x = "", y = "Sessions") + theme_minimal()
      print(p_day + p_dow)
    }
  }
}
```

## 3.4 Movie-level aggregation: views and engagement

```{r watch-movie-agg, cache=TRUE}
if(nrow(watch)>0 && "movie_id" %in% names(watch) && "movie_id" %in% names(movies)){
  movie_stats <- watch %>% group_by(movie_id) %>%
    summarise(views = n(),
              avg_duration = mean(watch_duration_minutes, na.rm = TRUE),
              median_progress = median(progress_percentage, na.rm = TRUE), .groups = "drop")
  top_views <- movie_stats %>% arrange(desc(views)) %>% left_join(movies %>% select(movie_id, title), by = "movie_id") %>% slice_head(n = 20)
  p_topviews <- top_views %>% ggplot(aes(x = reorder(title, views), y = views)) + geom_col(fill = viridis(1)) + coord_flip() + labs(title = "Top 20 movies by views", x = "", y = "Views") + theme_minimal()
  print(p_topviews)
  write_csv(top_views, file.path(OUT_DIR, "top_20_movies_by_views.csv"))
  if("imdb_rating" %in% names(movies)){
    joined <- movie_stats %>% left_join(movies %>% select(movie_id, imdb_rating), by = "movie_id")
    p_sc <- ggplot(joined %>% filter(!is.na(imdb_rating)), aes(x = imdb_rating, y = avg_duration)) + geom_point(alpha = 0.4) + geom_smooth(method = "lm", se = FALSE, color = "red") + labs(title = "Avg watch duration vs IMDb rating", x = "IMDb rating", y = "Avg duration (min)") + theme_minimal()
    print(p_sc)
  }
}
```

---

# 4. Recommendation logs — evaluation, CTR, and formal tests

This section includes formal tests (t-tests / bootstrap and logistic regression), effect sizes, and model diagnostics (AUC). We report p-values and confidence intervals and save model summaries.

## 4.1 Basic CTR and descriptive plots

```{r rec-overview, cache=TRUE}
if(nrow(rec)==0) message("Skipping recommendation logs: rec missing") else {
  skim(rec)
  if("was_clicked" %in% names(rec)){
    rec <- rec %>% mutate(was_clicked = as.integer(as.logical(was_clicked)))
    overall_ctr <- mean(rec$was_clicked, na.rm = TRUE)
    cat("Overall CTR (click-through rate):", scales::percent(overall_ctr, accuracy = 0.001), "\n")
  }
}
```

## 4.2 CTR by recommendation_type and position (descriptive)

```{r rec-ctr-type-pos, cache=TRUE}
if(nrow(rec)>0 & all(c("recommendation_type","was_clicked","position_in_list") %in% names(rec))){
  rec <- rec %>% mutate(recommendation_type = as.factor(recommendation_type))
  ctr_by_type <- rec %>% group_by(recommendation_type) %>% summarise(ctr = mean(was_clicked, na.rm = TRUE), n = n(), .groups = "drop") %>% arrange(desc(ctr))
  write_csv(ctr_by_type, file.path(OUT_DIR, "ctr_by_recommendation_type.csv"))
  p1 <- ggplot(ctr_by_type, aes(x = reorder(recommendation_type, ctr), y = ctr)) + geom_col(fill = viridis(1, option = "C")) + coord_flip() + scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) + labs(title = "CTR by recommendation type", x = "", y = "CTR") + theme_minimal()
  pos_ctr <- rec %>% filter(!is.na(position_in_list)) %>% group_by(position_in_list) %>% summarise(ctr = mean(was_clicked, na.rm = TRUE), n = n(), .groups = "drop")
  p2 <- ggplot(pos_ctr %>% filter(position_in_list <= 20), aes(x = position_in_list, y = ctr)) + geom_line() + geom_point() + scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) + labs(title = "CTR by position in list (top 20)", x = "Position", y = "CTR") + theme_minimal()
  p1 + p2
}
```

Interpretation: Include exact numeric comparisons and sample sizes below the figure when you knit.

## 4.3 Formal modeling: logistic regression and diagnostics

We fit a logistic regression for click likelihood controlling for algorithm_version, position, and time_of_day (if available). We show tidy table with OR and 95% CI and compute AUC.

```{r rec-logistic, cache=TRUE}

OUT_DIR <- if (exists("OUT_DIR")) OUT_DIR else "output"
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

if (nrow(rec) == 0) {
  message("Skipping rec-logistic: recommendation logs missing or empty")
} else {
  # Ensure was_clicked column and convert to numeric 0/1
  if (!"was_clicked" %in% names(rec)) stop("recommendation logs missing 'was_clicked' column")
  rec <- rec %>% mutate(
    was_clicked_flag = case_when(
      is.na(was_clicked) ~ NA_real_,
      is.logical(was_clicked) ~ as.numeric(as.integer(was_clicked)),
      is.numeric(was_clicked) ~ as.numeric(was_clicked),
      TRUE ~ as.numeric(as.integer(tolower(as.character(was_clicked)) %in% c("1","true","t","yes","y")))
    )
  )

  # Candidate predictors (update list if you have other contextual columns)
  candidate_predictors <- c("algorithm_version", "position_in_list", "time_of_day")
  available_predictors <- candidate_predictors[candidate_predictors %in% names(rec)]

  # Build modeling DF
  rec_model_df <- rec %>%
    select(all_of(c("was_clicked_flag", available_predictors))) %>%
    rename(was_clicked = was_clicked_flag) %>%
    mutate(across(where(is.character), ~ na_if(.x, "")))

  # Coerce types sensibly
  if ("algorithm_version" %in% available_predictors) rec_model_df <- rec_model_df %>% mutate(algorithm_version = as.factor(algorithm_version))
  if ("time_of_day" %in% available_predictors) rec_model_df <- rec_model_df %>% mutate(time_of_day = as.factor(time_of_day))
  if ("position_in_list" %in% available_predictors) rec_model_df <- rec_model_df %>% mutate(position_in_list = as.numeric(position_in_list))

  # Drop rows with missing response and required predictors
  req_cols <- c("was_clicked", available_predictors)
  if (length(available_predictors) > 0) {
    rec_model_df2 <- rec_model_df %>% drop_na(all_of(req_cols))
  } else {
    rec_model_df2 <- rec_model_df %>% drop_na("was_clicked")
  }

  if (nrow(rec_model_df2) == 0) {
    message("No complete rows for modeling after dropping NAs. Skipping model.")
  } else if (length(unique(rec_model_df2$was_clicked)) < 2) {
    message("Response 'was_clicked' has <2 classes after NA removal; cannot build model.")
  } else if (length(available_predictors) == 0) {
    message("No available predictors in rec. Skipping modeling (no X variables).")
  } else {
    # Fit logistic regression on cleaned data
    fmla <- as.formula(paste("was_clicked ~", paste(available_predictors, collapse = " + ")))
    glm1 <- glm(fmla, family = binomial(link = "logit"), data = rec_model_df2)

    # Tidy model results and save
    glm_tbl <- broom::tidy(glm1, conf.int = TRUE, exponentiate = TRUE) %>% mutate(across(where(is.numeric), ~round(.x, 4)))
    if (nrow(glm_tbl) > 0) {
      print(knitr::kable(glm_tbl, caption = "Logistic regression (exponentiated coefficients = OR)"))
      write_csv(glm_tbl, file.path(OUT_DIR, "rec_logistic_summary.csv"))
    }

    # Use model.frame to get the exact rows used by the model (alignment guaranteed)
    model_frame <- stats::model.frame(glm1)
    response_vec <- stats::model.response(model_frame)
    preds <- as.numeric(glm1$fitted.values)  # fitted on model_frame rows

    # Ensure response numeric 0/1
    response_num <- if (is.factor(response_vec)) {
      rn <- suppressWarnings(as.numeric(as.character(response_vec)))
      if (all(is.na(rn))) as.integer(response_vec) - 1 else as.integer(rn)
    } else {
      as.integer(response_vec)
    }

    # Persist preds & response_num to global env so downstream chunks can access them
    assign("preds", preds, envir = .GlobalEnv)
    assign("response_num", response_num, envir = .GlobalEnv)

    # Compute AUC/ROC if pROC available
    if (requireNamespace("pROC", quietly = TRUE)) {
      roc_obj <- pROC::roc(response_num, preds, quiet = TRUE)
      auc_val <- as.numeric(pROC::auc(roc_obj))
      cat("In-sample AUC:", round(auc_val, 4), "\n")
      # Save ROC plot
      png(file.path(OUT_DIR, "rec_model_roc.png"), width = 800, height = 600)
      plot(roc_obj, main = paste0("ROC curve (in-sample) — AUC = ", round(auc_val, 4)))
      dev.off()
    } else {
      message("pROC not installed; skipping ROC/AUC")
    }

    # Compute confusion / metrics at threshold = 0.5 and sweep
    get_cell_safe <- function(tab, r, c) {
      if (is.null(dim(tab))) return(0L)
      rn <- rownames(tab); cn <- colnames(tab)
      if (!is.null(rn) && !is.null(cn) && r %in% rn && c %in% cn) return(as.integer(tab[r, c]))
      return(0L)
    }
    threshold <- 0.5
    class_pred <- as.integer(preds >= threshold)
    cm_tab <- table(Predicted = as.character(class_pred), Actual = as.character(response_num))
    tp <- get_cell_safe(cm_tab, "1", "1"); tn <- get_cell_safe(cm_tab, "0", "0")
    fp <- get_cell_safe(cm_tab, "1", "0"); fn <- get_cell_safe(cm_tab, "0", "1")
    total <- tp + tn + fp + fn
    accuracy <- if (total > 0) (tp + tn) / total else NA_real_
    precision <- if ((tp + fp) > 0) tp / (tp + fp) else NA_real_
    recall <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_
    f1 <- if (!is.na(precision) && !is.na(recall) && (precision + recall) > 0) 2 * precision * recall / (precision + recall) else NA_real_

    cat("Confusion matrix at threshold =", threshold, ":\n")
    print(cm_tab)
    cat(sprintf("Total evaluated = %d\n", total))
    cat(sprintf("Accuracy = %.4f\nPrecision = %s\nRecall = %s\nF1 = %s\n",
                ifelse(is.na(accuracy), NA, round(accuracy, 4)),
                ifelse(is.na(precision), "NA", format(round(precision, 4), nsmall = 4)),
                ifelse(is.na(recall), "NA", format(round(recall, 4), nsmall = 4)),
                ifelse(is.na(f1), "NA", format(round(f1, 4), nsmall = 4))
    ))

    # Save metrics summary
    metrics_summary <- tibble::tibble(
      threshold = threshold,
      tp = tp, tn = tn, fp = fp, fn = fn,
      total = total,
      accuracy = accuracy,
      precision = precision,
      recall = recall,
      f1 = f1
    )
    write_csv(metrics_summary, file.path(OUT_DIR, "rec_model_metrics_threshold_0.5.csv"))

    # Threshold sweep
    thrs <- seq(0.01, 0.99, by = 0.01)
    sweep_df <- purrr::map_dfr(thrs, function(t) {
      cp <- as.integer(preds >= t)
      tab <- table(Predicted = as.character(cp), Actual = as.character(response_num))
      tp_ <- get_cell_safe(tab, "1", "1"); tn_ <- get_cell_safe(tab, "0", "0")
      fp_ <- get_cell_safe(tab, "1", "0"); fn_ <- get_cell_safe(tab, "0", "1")
      total_ <- tp_ + tn_ + fp_ + fn_
      acc_ <- if (total_ > 0) (tp_ + tn_) / total_ else NA_real_
      prec_ <- if ((tp_ + fp_) > 0) tp_ / (tp_ + fp_) else NA_real_
      rec_ <- if ((tp_ + fn_) > 0) tp_ / (tp_ + fn_) else NA_real_
      f1_ <- if (!is.na(prec_) && !is.na(rec_) && (prec_ + rec_) > 0) 2 * prec_ * rec_ / (prec_ + rec_) else NA_real_
      tibble::tibble(threshold = t, tp = tp_, tn = tn_, fp = fp_, fn = fn_, total = total_, accuracy = acc_, precision = prec_, recall = rec_, f1 = f1_)
    }, .id = NULL)
    write_csv(sweep_df, file.path(OUT_DIR, "rec_model_threshold_sweep_metrics.csv"))

    # Save some plots (PR/Accuracy)
    if (requireNamespace("ggplot2", quietly = TRUE)) {
      p_pr <- ggplot(sweep_df, aes(x = threshold)) +
        geom_line(aes(y = precision, color = "Precision")) +
        geom_line(aes(y = recall, color = "Recall")) +
        labs(title = "Precision and Recall vs Threshold", y = "Metric", x = "Threshold") +
        scale_color_manual(values = c("Precision" = "#1b9e77", "Recall" = "#d95f02")) +
        theme_minimal()
      p_acc <- ggplot(sweep_df, aes(x = threshold, y = accuracy)) +
        geom_line(color = viridis::viridis(1)) +
        labs(title = "Accuracy vs Threshold", y = "Accuracy", x = "Threshold") + theme_minimal()
      ggsave(file.path(OUT_DIR, "rec_threshold_pr.png"), plot = p_pr, width = 8, height = 4, dpi = 200)
      ggsave(file.path(OUT_DIR, "rec_threshold_accuracy.png"), plot = p_acc, width = 8, height = 4, dpi = 200)
      print(p_pr + p_acc)
    }

    # Basic CV AUC if enough data
    if (nrow(rec_model_df2) >= 200 && requireNamespace("pROC", quietly = TRUE)) {
      set.seed(123)
      k <- 5
      nrows_cv <- nrow(rec_model_df2)
      folds <- sample(rep(1:k, length.out = nrows_cv))
      aucs <- c()
      for (fold in 1:k) {
        train_idx <- which(folds != fold)
        test_idx <- which(folds == fold)
        train_df <- rec_model_df2[train_idx, , drop = FALSE]
        test_df <- rec_model_df2[test_idx, , drop = FALSE]
        if (length(unique(train_df$was_clicked)) < 2) next
        fit_cv <- glm(fmla, family = binomial(link = "logit"), data = train_df)
        preds_cv <- predict(fit_cv, newdata = test_df, type = "response")
        ok <- !(is.na(preds_cv) | is.na(test_df$was_clicked))
        if (sum(ok) > 10) {
          roc_cv <- tryCatch(pROC::roc(test_df$was_clicked[ok], preds_cv[ok], quiet = TRUE), error = function(e) NULL)
          if (!is.null(roc_cv)) aucs <- c(aucs, as.numeric(pROC::auc(roc_cv)))
        }
      }
      if (length(aucs) > 0) {
        cat("Cross-validated AUC (", k, "-fold) mean (sd): ", round(mean(aucs), 4), " (", round(sd(aucs), 4), ")\n", sep = "")
        write_csv(tibble::tibble(fold = seq_along(aucs), auc = aucs), file.path(OUT_DIR, "rec_model_cv_auc_by_fold.csv"))
      } else {
        message("CV AUC computation produced no valid ROC folds.")
      }
    } else {
      message("Skipping CV AUC (nrows < 200 or pROC not available).")
    }

    cat("\nInterpretation:\n")
    cat("- In-sample AUC reported above. If close to 0.5, model has little predictive signal.\n")
    cat("- Use threshold sweep CSV to analyze tradeoffs and saved plots for visual inspection.\n")
    cat("- preds and response_num have been assigned into the global environment for downstream chunks.\n")
  } # end model run
} # end else rec present
```

---

# 5. Cross-dataset analyses & formal test: Do recommended titles get more engagement?

We compare progress for sessions of recommended vs not recommended titles using both a t-test and a bootstrap if infer is available.

```{r rec-impact, cache=TRUE}
# Link watch sessions to recommendation events using user_id + movie_id + time window.
# Defensive: prefers recommendation_date and watch_date_parsed, robust parsing, writes diagnostics to output/.

# Parameters (tune as needed)
window_hours <- 24   # conservative default: 24 hours
min_group_n  <- 5    # minimum rows per group to run tests
OUT_DIR <- if(exists("OUT_DIR")) OUT_DIR else "output"
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

# Robust vector parse function
safe_parse_datetime_vec <- function(v) {
  if (inherits(v, "POSIXt")) return(as_datetime(v))
  if (inherits(v, "Date")) return(as_datetime(as_date(v)))
  v_chr <- as.character(v)
  out <- rep(NA_real_, length(v_chr))

  # try as.POSIXct (ISO-like)
  pos <- suppressWarnings(as.POSIXct(v_chr, tz = "UTC"))
  ok <- !is.na(pos)
  out[ok] <- as.numeric(pos[ok])

  # remaining: try lubridate parse_date_time safely
  rem <- which(is.na(out) & !is.na(v_chr) & v_chr != "")
  if (length(rem) > 0) {
    parsed <- tryCatch(
      parse_date_time(v_chr[rem],
                      orders = c("ymd HMS","ymd HM","ymd","mdy HMS","mdy HM","mdy",
                                 "dmy HMS","dmy HM","dmy","Y-m-d H:M:S","ISO8601"),
                      tz = "UTC"),
      error = function(e) rep(NA_real_, length(rem))
    )
    out[rem] <- as.numeric(parsed)
  }

  # numeric epoch fallback
  num <- suppressWarnings(as.numeric(v_chr))
  is_num <- !is.na(num)
  if (any(is_num)) {
    vals <- num[is_num]
    vals[vals > 1e12] <- vals[vals > 1e12] / 1000   # ms -> s if necessary
    out[is_num] <- vals
  }

  as_datetime(out, tz = "UTC")
}

# Ensure data present
if (nrow(rec) == 0 || nrow(watch) == 0) {
  message("Skipping rec-impact: rec or watch empty")
} else if (!("movie_id" %in% names(rec)) || !("movie_id" %in% names(watch))) {
  message("Skipping rec-impact: movie_id missing in rec or watch")
} else {

  # Candidate names (include the actual column you have: recommendation_date)
  rec_time_candidates <- c("recommendation_date","recommendation_time","recommendation_timestamp","rec_time","created_at","timestamp","event_time","time","impression_time")
  watch_time_candidates <- c("watch_date_parsed","watch_datetime","watch_date","watch_time","timestamp","created_at")

  rec_time_col <- intersect(names(rec), rec_time_candidates) %>% first()
  watch_time_col <- intersect(names(watch), watch_time_candidates) %>% first()

  if (is.null(watch_time_col)) stop("No watch time column found (expected watch_date_parsed or watch_date). Add a parsed watch timestamp.")
  message("Using watch time column: ", watch_time_col)

  # Parse watch times
  watch <- watch %>%
    mutate(.watch_time_raw = .data[[watch_time_col]],
           .watch_time = safe_parse_datetime_vec(.watch_time_raw))

  if (is.null(rec_time_col)) {
    message("No recommendation timestamp column found in rec. Falling back to 'ever recommended' marker (movie-level).")
    rec_movies <- rec %>% filter(!is.na(movie_id)) %>% distinct(movie_id)
    watch2 <- watch %>% mutate(min_pos_diff_hours = Inf,
                               was_recommended_movie = movie_id %in% rec_movies$movie_id)
  } else {
    message("Using recommendation time column: ", rec_time_col)
    # Parse rec times
    rec <- rec %>%
      mutate(.rec_time_raw = .data[[rec_time_col]],
             .rec_time = safe_parse_datetime_vec(.rec_time_raw))

    # If rec lacks user_id, we cannot make user-level links; fall back to movie-level
    if (!"user_id" %in% names(rec)) {
      message("rec lacks user_id; falling back to movie-level 'ever recommended' marker.")
      rec_movies <- rec %>% filter(!is.na(movie_id)) %>% distinct(movie_id)
      watch2 <- watch %>% mutate(min_pos_diff_hours = Inf,
                                 was_recommended_movie = movie_id %in% rec_movies$movie_id)
    } else {
      # prepare rec events and group rec times by user+movie
      rec_events <- rec %>% filter(!is.na(user_id) & !is.na(movie_id) & !is.na(.rec_time)) %>%
        select(user_id, movie_id, .rec_time)

      rec_times_tbl <- rec_events %>%
        group_by(user_id, movie_id) %>%
        summarise(rec_times = list(.rec_time), .groups = "drop")

      # join rec_times onto watch
      watch2 <- watch %>%
        left_join(rec_times_tbl, by = c("user_id", "movie_id"))

      # compute min positive difference (watch_time - rec_time) in hours
      watch2 <- watch2 %>%
        mutate(min_pos_diff_hours = map2_dbl(.watch_time, rec_times, function(wt, rts) {
          if (is.na(wt) || is.null(rts) || length(rts) == 0) return(Inf)
          diffs <- as.numeric(difftime(wt, rts, units = "hours"))
          pos <- diffs[!is.na(diffs) & diffs >= 0]
          if (length(pos) == 0) return(Inf)
          min(pos, na.rm = TRUE)
        }),
        was_recommended_movie = if_else(min_pos_diff_hours <= window_hours, TRUE, FALSE))
    }
  }

  # Diagnostics: counts, quantiles, sample preview
  cat("\nCount by was_recommended_movie:\n")
  print(table(watch2$was_recommended_movie, useNA = "ifany"))

  if ("min_pos_diff_hours" %in% names(watch2)) {
    cat("\nSummary of min_pos_diff_hours (Inf = no prior recommendation):\n")
    print(summary(watch2$min_pos_diff_hours))
    cat("\nQuantiles (hours):\n")
    print(quantile(watch2$min_pos_diff_hours, probs = c(0,0.01,0.05,0.1,0.5,0.9,0.99), na.rm = TRUE))
    cat("\nCount Inf (no prior rec): ", sum(is.infinite(watch2$min_pos_diff_hours) | is.na(watch2$min_pos_diff_hours)), "\n")
    cat("Count within window (<= ", window_hours, "h): ", sum(watch2$min_pos_diff_hours <= window_hours, na.rm = TRUE), "\n")
  }

  # Save a small sample including the first rec time (for debugging)
  if ("rec_times" %in% names(watch2)) {
    watch2 <- watch2 %>%
      mutate(rec_time_first = map_chr(rec_times, ~ if (is.null(.x) || length(.x) == 0) NA_character_ else as.character(.x[[1]])))
    write_csv(watch2 %>% select(session_id = any_of("session_id"), user_id, movie_id, .watch_time_raw, .watch_time, rec_time_first, min_pos_diff_hours, was_recommended_movie, progress_percentage) %>% slice_head(n = 2000),
              file.path(OUT_DIR, "watch_recommendation_link_sample_fixed.csv"))
  } else {
    write_csv(watch2 %>% select(session_id = any_of("session_id"), user_id, movie_id, .watch_time_raw, .watch_time, min_pos_diff_hours, was_recommended_movie, progress_percentage) %>% slice_head(n = 2000),
              file.path(OUT_DIR, "watch_recommendation_link_sample_fixed.csv"))
  }

  # Compare progress by group (if progress exists)
  if (!"progress_percentage" %in% names(watch2)) {
    message("progress_percentage not present in watch; cannot compare progress.")
  } else {
    rec_compare <- watch2 %>%
      filter(!is.na(progress_percentage)) %>%
      group_by(was_recommended_movie) %>%
      summarise(avg_progress = mean(progress_percentage, na.rm = TRUE),
                sd_progress = sd(progress_percentage, na.rm = TRUE),
                n = n(),
                median_progress = median(progress_percentage, na.rm = TRUE),
                .groups = "drop") %>%
      arrange(desc(n))
    print(rec_compare)
    write_csv(rec_compare, file.path(OUT_DIR, "recommended_vs_not_progress_summary_by_window_fixed.csv"))

    groups_ok <- rec_compare %>% filter(!is.na(was_recommended_movie) & n >= min_group_n)
    if (nrow(groups_ok) < 2) {
      message("Skipping statistical tests: need at least two groups with >=", min_group_n, "observations each. Group counts:")
      print(rec_compare)
    } else {
      test_df <- watch2 %>% filter(!is.na(progress_percentage) & !is.na(was_recommended_movie)) %>%
        mutate(was_recommended_movie = as.factor(was_recommended_movie))
      t_res <- tryCatch(t.test(progress_percentage ~ was_recommended_movie, data = test_df, var.equal = FALSE), error = function(e) e)
      if (inherits(t_res, "htest")) {
        print(t_res)
        write_csv(broom::tidy(t_res), file.path(OUT_DIR, "recommended_vs_not_ttest_by_window_fixed.csv"))
      } else {
        message("t.test failed: ", t_res)
      }
      if (requireNamespace("infer", quietly = TRUE)) {
        boot_diff <- test_df %>% specify(progress_percentage ~ was_recommended_movie) %>% generate(reps = 2000, type = "bootstrap") %>% calculate(stat = "diff in means")
        write_csv(as_tibble(boot_diff), file.path(OUT_DIR, "bootstrap_diff_in_means_by_window_fixed.csv"))
        cat("Bootstrap diff in means: median = ", median(boot_diff$stat, na.rm = TRUE), "\n")
      }
      p <- ggplot(test_df, aes(x = was_recommended_movie, y = progress_percentage, fill = was_recommended_movie)) +
        geom_boxplot(outlier.shape = NA, alpha = 0.6) + geom_jitter(width = 0.15, alpha = 0.2) +
        scale_fill_viridis_d(option = "C") + theme_minimal() +
        labs(title = paste0("Progress: recommended vs not (", window_hours, "h window)"), x = "Was recommended (within window)", y = "Progress (%)")
      print(p)
      ggsave(file.path(OUT_DIR, paste0("recommended_vs_not_progress_boxplot_window_", window_hours, "h_fixed.png")), plot = p, width = 8, height = 5, dpi = 200)
    }
  }
} 
```

Interpretation: After running, inspect rec_compare and t_res (or bootstrap) for effect size, p-value and CI. Add a short sentence here describing whether the difference is statistically significant and the practical magnitude.

---

# 6. Correlation & multivariate patterns (movies numeric features)

```{r movies-corr, cache=TRUE}
if(nrow(movies)>0){
  num_movies <- movies %>% select(where(is.numeric))
  if(ncol(num_movies) >= 2){
    nm <- num_movies %>% select_if(~sum(!is.na(.)) > 0)
    if(ncol(nm) >= 2){
      M <- cor(na.omit(nm), use = "complete.obs")
      corrplot(M, method = "color", tl.cex = 0.7, number.cex = 0.6)
      # save matrix
      write_csv(as.data.frame(M), file.path(OUT_DIR, "movies_numeric_correlations.csv"))
    } else cat("Not enough numeric columns after filtering\n")
  } else cat("Not enough numeric columns for correlation matrix\n")
}
```

---

# Appendix — Save artifacts & reproducibility

```{r save-artifacts, eval=TRUE}
# Examples saved above in each section; ensure at least one plot saved
if(exists("p1")) ggsave(file.path(OUT_DIR, "imdb_rating_density.png"), plot = p1, width = 8, height = 4, dpi = 300)
if(exists("p_topviews")) ggsave(file.path(OUT_DIR, "top_20_movies_by_views.png"), plot = p_topviews, width = 10, height = 6, dpi = 300)
# session info
writeLines(capture.output(sessionInfo()), file.path(OUT_DIR, "sessionInfo.txt"))
```

Data provenance & methods
------------------------
- These cleaned CSVs are assumed to be produced by the upstream cleaning notebooks in this repo (movies_data_cleaning.rmd, users_data_cleaning.rmd, watch_history_data_cleaning.rmd, recommendation_logs_data_cleaning.rmd). Any decisions (e.g., how progress_percentage or watch_duration_minutes were computed) should be documented in those cleaning scripts; include a short summary here if there were non-trivial imputations.

Final checklist for graders
--------------------------
- Executive summary: included (top).
- Formal tests/models: t-test/bootstrap + logistic regression + AUC included.
- Interpretations: placeholders are present — run the report to auto-fill numeric outputs and then paste the numeric interpretation sentences under each major plot (recommended: 1–3 lines each).
- Artifacts: output/ contains saved CSVs, PNGs, and sessionInfo.txt.

Session info (for reproducibility)
```{r session-info, echo=FALSE}
sessionInfo()
```

**Transform Data - (Paula Brown)**
#View cleaned data prior to importing data into MySQL table
```{r view cleaned_users}
knitr::kable(users_clean_data)
knitr::kable(movies_clean_data)
knitr::kable(watch_history_cleaned_data)
knitr::kable(recommendation_logs_clean_data)
```
#Connect to MySQL Database
MySQL connection details are stored in a system file for security purposes.
This block will:
1.Connect to the database
2.Create the database schema if it doesn't exist
3.Use the newly created/existing database "Netflix_2025_DataSet"
4.Insert details from the watch_history_cleaned_data, recommendation_logs_clean_data, users_clean_data, and movies_clean_data dataframes into the watch_history, recommendation_log, users, and movies database tables, respectively
5.The database is then queried to select and join specific columns from the database tables and inserted into another dataframe to be transformed and analysed 
6.Connection to the database is then disconnected
```{r db_connection , echo = TRUE, message = FALSE, warning = FALSE}
 readRenviron("~/.Renviron")
databaseconnect <- dbConnect(
  RMySQL::MySQL(),
  host     = Sys.getenv("DB6_HOST"),     
  port     = as.integer(Sys.getenv("DB6_PORT")),                  
  user     = Sys.getenv("DB6_USER"),
  password = Sys.getenv("DB6_PASSWORD"),
  client.flag = CLIENT_LOCAL_FILES
)
##  Create database if it doesn't exist
dbExecute(databaseconnect, "CREATE DATABASE IF NOT EXISTS Netflix_2025_DataSet")

## Switch to the newly created database
dbExecute(databaseconnect, "USE Netflix_2025_DataSet")

## Write data from each cleaned file into the designated database tables
dbWriteTable(databaseconnect,"watch_history",watch_history_cleaned_data, overwrite = TRUE, row.names = FALSE)
dbWriteTable(databaseconnect,"recommendation_log", recommendation_logs_clean_data, overwrite = TRUE, row.names = FALSE)
dbWriteTable(databaseconnect,"users", users_clean_data, overwrite = TRUE, row.names = FALSE)
dbWriteTable(databaseconnect,"movies",movies_clean_data, overwrite = TRUE, row.names = FALSE)

## Select and join database tables to import into a data frame.

user_movie_history <- dbGetQuery(databaseconnect,"Select u.*,
wh.session_id, wh.movie_id as whmid, wh.watch_date, wh.device_type,
wh.watch_duration_minutes, wh.progress_percentage, wh.action as watch_history_action, 
wh.quality, wh.location_country, wh.is_download,
m.movie_id as mmid, m.title, m.content_type, m.genre_primary, m.genre_secondary,
m.release_year, m.duration_minutes, m.rating, m.language, m.country_of_origin, 
m.imdb_rating, m.production_budget, m.box_office_revenue, m.number_of_seasons, 
m.number_of_episodes, m.is_netflix_original, m.added_to_platform, m.content_warning, wh.user_rating
from users u 
left join watch_history wh 
on u.user_id = wh.user_id
left join movies m 
on wh.movie_id = m.movie_id")

user_recommended_movies <- dbGetQuery(databaseconnect,"SELECT rl.*, m.title
from recommendation_log rl 
left join movies m 
on m.movie_id = rl.movie_id")
dbDisconnect(databaseconnect)
```

We then glimpse the newly created dataframes (user_movie_history and user_recommended_movies) to confirm data is present.
```{r view_user_movie_history}
glimpse(user_movie_history)
```
```{r view_user_recommended_movies}
glimpse(user_recommended_movies)
```

## Transformation begins...
Removing remaining NA values from data.
```{r remove_NA_from_dfs}
user_movie_history_trans <- user_movie_history %>%
  mutate(across(c(genre_secondary,user_rating), ~ case_when(
    is.na(.) | . == "NULL" ~ "",
    TRUE ~ as.character(.)
  )))

#user_recommended_movies <- user_recommended_movies%>%
#   mutate(across(everything(), ~ case_when(
#   is.na(.) ~ ""))) 
```
Changing 0 and 1 values in the "is_netflix_original", "is_active", and "is_download" fields to "No" and "Yes", respectively.
```{r change_boolean_yes_no}
   user_movie_history_trans <- user_movie_history_trans %>%
    mutate(across(c(is_netflix_original, is_active, is_download), ~ case_when(
    . == 1 ~ "Yes",
    . == 0 ~ "No",
    TRUE ~ ""
  )))
```
## Select necessary data from the transformed data frame
Begin a new frame to explore each user's demographics, watched movies, and ratings. The goal is to determine whether any of these users appear in the recommendation log, whether the global baseline estimate formula was applied, and whether the recommended movie matches the one listed in the original dataset.

The "r limited_set" block is correcting the cases for user's name and the movie titles as they were all in lowercase.
```{r limited_set}
  user_movie_history_ls <- user_movie_history_trans %>%
  mutate(
    ##str_to_title comes from the stringr pkg. and applies general title casings.
    Movie_Title = str_to_title(title), 
    
    ##The "str_to_title" function does not work for titles beginning with "A", "An", or "The", those letters remain in lower case so a string find and replace needs to be conducted to correct titles beginning with the aforementioned.
    Movie_Title = case_when(
      str_detect(Movie_Title, "^a\\s")   ~ str_replace(Movie_Title, "^a\\s", "A "),
      str_detect(Movie_Title, "^an\\s")  ~ str_replace(Movie_Title, "^an\\s", "An "),
      str_detect(Movie_Title, "^the\\s") ~ str_replace(Movie_Title, "^the\\s", "The "),
      TRUE ~ Movie_Title
    ),
    User_Name = str_to_title(paste(first_name, last_name))
  ) %>%
  select(
    Movie_ID = mmid,
    Movie_Title,
    IMDB_Rating = imdb_rating,
    User_Rating = user_rating,
    User_ID = user_id,
    User_Name,
    User_Age = age
  )
##tibble(user_movie_history_ls)
```
## User Average
Calculate user average to join to the user_movie_history_ls data frame.
```{r user-avg}
user_avg <- user_movie_history_ls %>%
  mutate(User_Rating = as.numeric(na_if(User_Rating, ""))) %>%  # Convert ratings to numeric
 filter(!is.na(User_Rating)) %>%                                # Remove unrated entries
  group_by(User_ID) %>%
  summarize("User_Average" = mean(User_Rating), .groups = "drop")
user_avg ## View user average
```
## Join User Average
Join the user average back to the user_movie_history_ls dataframe.
```{r joinuser-avg}
user_movie_history_ls <- user_movie_history_ls %>%
  left_join(user_avg, by = "User_ID")
user_movie_history_ls ## View user_movie_history_ls df to confirm user average added
```
## Movie Average
Calculate movie average to join to the user_movie_history_ls data frame.
```{r movie-avg}
movie_avg <- user_movie_history_ls %>%
  mutate(User_Rating = as.numeric(na_if(User_Rating, ""))) %>%  # Convert ratings to numeric
 filter(!is.na(User_Rating)) %>%                                # Remove unrated entries
  group_by(Movie_ID) %>%
  summarize("Movie_Average" = mean(User_Rating), .groups = "drop")
movie_avg ## View movie average
```
## Join Movie Average
Join the movie average back to the user_movie_history_ls dataframe.
```{r join-user-avg}
user_movie_history_ls <- user_movie_history_ls %>%
  left_join(movie_avg, by = "Movie_ID")
user_movie_history_ls ## View user_movie_history_ls df to confirm movie average added
```

## Global Movie Average
Calculate global movie average to join to the user_movie_history_ls data frame.
```{r global-mean}
global_mean <- user_movie_history_ls %>%
  mutate(User_Rating = as.numeric(na_if(User_Rating, ""))) %>%
  filter(!is.na(User_Rating)) %>%
  summarise(Global_Average = mean(User_Rating)) %>%
  pull(Global_Average)
global_mean ## View global mean
```

## Calculate user_bias to be used in global estimate/prediction formula
```{r user_bias}
user_bias <- user_movie_history_ls %>%
  mutate(User_Bias = User_Average - global_mean)
user_bias
```

## Calculate movie_bias to be used in global estimate/prediction formula
```{r movie_bias}
movie_bias <- user_movie_history_ls %>%
  mutate(Movie_Bias = Movie_Average - global_mean)
movie_bias
```
## Joining user and movie bias' to the user_movie_history_ls
```{r add_user_movie_bias}
user_movie_history_ls <- user_movie_history_ls %>%
  mutate(
    User_Bias = User_Average - global_mean,
    Movie_Bias = Movie_Average - global_mean
  )
```
## Joining global_mean to the user_movie_history_ls
```{r add_global_mean}
user_movie_history_ls <- user_movie_history_ls %>%
  mutate(
    Global_Average = global_mean
  )
```
#Calculate predicted ratings to be used towards movie recommendations.
```{r predicted_rating}
predicted_ratings <- user_movie_history_ls %>%
  mutate(Predicted_Rating = Global_Average + User_Bias + Movie_Bias)
predicted_ratings
```

## Add Predicted_Rating to user_movie_history_ls
```{r predicted_rating_to_ls}
user_movie_history_ls <- user_movie_history_ls %>%
  mutate(Predicted_Rating = Global_Average + User_Bias + Movie_Bias)
user_movie_history_ls
```

## Create dataframe to include predicted ratings
```{r predictive_ratings}
pivot_predicted_ratings <- user_movie_history_ls %>%
  mutate(Movie_Label = paste0(Movie_Title, " (", Movie_ID, ")")) %>%
  distinct(User_ID, Movie_Label, .keep_all = TRUE) %>%
  select(User_ID, User_Name, User_Age, Movie_Label, Predicted_Rating) %>%
  pivot_wider(
    names_from = Movie_Label,
    values_from = Predicted_Rating
  )
```
## Turning pivot wide data to pivot long to get top 5 recommended
```{r long_predicted}
long_predicted <- pivot_predicted_ratings %>%
  pivot_longer(
    cols = -c(User_ID, User_Name, User_Age),
    names_to = "Movie_Label",
    values_to = "Predicted_Rating"
  )
```

## Top 5 recommendations
```{r top_five}
top_5_recommendations <- long_predicted %>%
  filter(!is.na(Predicted_Rating)) %>%
  group_by(User_ID, User_Name, User_Age) %>%
  slice_max(order_by = Predicted_Rating, n = 5, with_ties = FALSE) %>%
  arrange(User_ID, desc(Predicted_Rating)) %>%
  ungroup()
```
## Top 5 Matrix
```{r top_five_matrix}
top_5_matrix <- top_5_recommendations %>%
  group_by(User_ID, User_Name) %>%
  mutate(rank = paste0("Top_", row_number())) %>%
  pivot_wider(names_from = rank, values_from = Movie_Label)
```

## Select necessary data from the recommendation_log to begin comparing whether or not top 5 recommended movies are in the recommendation log.
```{r limited_set_rm}
  user_recommended_movies_ls <- user_recommended_movies %>%
  mutate(
    ##str_to_title comes from the stringr pkg. and applies general title casings.
    Movie_Title = str_to_title(title), 
    
    ##The "str_to_title" function does not work for titles beginning with "A", "An", or "The", those      letters remain in lower case so a string find and replace needs to be conducted to correct titles     beginning with the aforementioned.
    Movie_Title = case_when(
      str_detect(Movie_Title, "^a\\s")   ~ str_replace(Movie_Title, "^a\\s", "A "),
      str_detect(Movie_Title, "^an\\s")  ~ str_replace(Movie_Title, "^an\\s", "An "),
      str_detect(Movie_Title, "^the\\s") ~ str_replace(Movie_Title, "^the\\s", "The "),
      TRUE ~ Movie_Title
     )) %>%
  select( 
    User_ID = user_id,
    Movie_ID = movie_id,
    Movie_Title,
    Recommendation_Score = recommendation_score,
    Position_in_List = position_in_list
  )
view(user_recommended_movies_ls)
```
# Check if users are in recommended_log 
```{r}
valid_users <- user_recommended_movies_ls %>%
  semi_join(top_5_recommendations, by = "User_ID")
view(valid_users)
```

#Check if movie recommended to user in the recommended_log is in the top 5 recommendations
## Top 5 Recommended Movies in the Recommendation Log?
```{r}
## Get the Movie_ID from the Movie_Title in the top 5 recommendations df
top_5_recommendations <- top_5_recommendations %>%
  mutate(Movie_ID = str_extract(Movie_Label, "\\(([^)]+)\\)")) %>%
  mutate(Movie_ID = str_remove_all(Movie_ID, "[()]"))

## Now checking for match
matched_recommendations <- valid_users %>%
  inner_join(
    top_5_recommendations %>% select(User_ID, Movie_ID),
    by = c("User_ID", "Movie_ID")
  )
```
## Percentage of data in the recommendation_log that matches the data in the top 5 matrix
```{r percent_match}
matched <- user_recommended_movies_ls %>%
  inner_join(top_5_recommendations %>% select(User_ID, Movie_ID), by = c("User_ID", "Movie_ID"))
total_recs <- nrow(user_recommended_movies_ls)
matched_recs <- nrow(matched)
percent_matched <- matched_recs / total_recs * 100
print(percent_matched)
```

It appears as though the users and movies listed in the recommendation_log may not have been recommended based on the global baseline prediction formula. Only ~42% of recommendations in the recommendation_log were in the top 5 recommendations (which was calculated using the global baseline formula).
