---
title: "Project3"
Author: "Mehreen ALi Gillani"
output: html_document
date: "2025-10-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

####  Import Libraries

```{r import libraries}
library(tidyverse)

```
### Read users, movies, watch_history and recommendation_logs csv file put them in Dataframes

```{r}
users = read.csv('https://raw.githubusercontent.com/mehreengillani/DATA607_project3/refs/heads/main/users_clean.csv')

watch_history = read.csv('https://raw.githubusercontent.com/mehreengillani/DATA607_project3/refs/heads/main/watch_history_clean.csv')

movies = read.csv('https://raw.githubusercontent.com/mehreengillani/DATA607_project3/refs/heads/main/movies_clean.csv')

recommendation_logs = read.csv('https://raw.githubusercontent.com/mehreengillani/DATA607_project3/refs/heads/main/recommendation_logs_clean.csv')

```
### 1. Content Engagement Analysis

####  Question: Which subscription plans have the highest user engagement (watch time per user) and does engagement correlate with monthly spend?

```{r}
engagement_by_plan <- watch_history %>%
  group_by(user_id) %>%
  summarize(total_watch_time = sum(watch_duration_minutes, na.rm = TRUE)) %>%
  inner_join(users %>% select(user_id, subscription_plan, monthly_spend), by = "user_id") %>%
  group_by(subscription_plan) %>%
  summarize(
    avg_watch_time_per_user = mean(total_watch_time),
    avg_monthly_spend = mean(monthly_spend),
    user_count = n()
  )

engagement_by_plan
```
```{r}
library(ggplot2)
library(dplyr)
library(scales)

# Prepare data
engagement_plot <- watch_history %>%
  group_by(user_id) %>%
  summarize(total_watch_time = sum(watch_duration_minutes, na.rm = TRUE)) %>%
  inner_join(users %>% select(user_id, subscription_plan, monthly_spend), by = "user_id") %>%
  group_by(subscription_plan) %>%
  summarize(
    avg_watch_hours = mean(total_watch_time) / 60,
    avg_monthly_spend = mean(monthly_spend),
    user_count = n()
  )
engagement_plot

# Plot
p1 <- ggplot(engagement_plot, aes(x = subscription_plan, y = avg_watch_hours, fill = subscription_plan)) +
  geom_col(alpha = 0.7) +
  geom_text(aes(label = paste0(round(avg_watch_hours, 1), " hrs")), vjust = 0.5) +
  labs(title = "Average Watch Time by Subscription Plan",
       subtitle = "Higher-tier plans show increased engagement",
       x = "Subscription Plan", y = "Average Watch Hours per User") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()

p2 <- ggplot(engagement_plot, aes(x = avg_monthly_spend, y = avg_watch_hours, 
                                 color = subscription_plan, size = user_count)) +
  geom_point(alpha = 0.7) +
  geom_text(aes(label = subscription_plan), vjust = -1, size = 5) +
  labs(title = "Monthly Spend vs Watch Time",
       x = "Average Monthly Spend ($)", y = "Average Watch Hours",
       size = "Number of Users") +
  theme_minimal()

p1 / p2  # Using patchwork for layout
```

####  key findings:

Minimal engagement differences - All plans show nearly identical watch times (10.74-10.93 hours) despite different price points

Standard plan offers poorest value - Highest monthly cost ($16.24) with mid-range viewing time

Premium+ provides best value - Second-lowest cost with comparable engagement to more expensive plans

### 2. Content Performance vs. Production Cost

####  Question: Do higher production budgets lead to better user ratings and engagement for Netflix Originals?

```{r}
# Prepare data
budget_performance <- movies %>%
  filter(is_netflix_original == TRUE & 
         !is.na(production_budget) & 
         production_budget > 0) %>%  # Added filter for non-zero budgets
  inner_join(
    watch_history %>% 
      group_by(movie_id) %>%
      summarize(
        avg_user_rating = mean(user_rating, na.rm = TRUE),
        total_watch_hours = sum(watch_duration_minutes, na.rm = TRUE) / 60
      ),
    by = "movie_id"
  ) %>%
  select(
    movie_id,
    title,
    production_budget,
    genre_primary,
    release_year,
    avg_user_rating,
    total_watch_hours
  )
# Plot showing both metrics against budget
p3 <- ggplot(budget_performance, aes(x = production_budget/1e6)) +
  geom_point(aes(y = avg_user_rating, size = total_watch_hours, color = "User Rating"), alpha = 0.7) +
  geom_smooth(aes(y = avg_user_rating), method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  geom_point(aes(y = total_watch_hours/30, color = "Watch Hours"), alpha = 0.7) +  # Scaled for visibility
  geom_smooth(aes(y = total_watch_hours/30), method = "lm", se = FALSE, color = "blue", linetype = "dashed") +
  scale_x_continuous(labels = dollar_format(prefix = "$", suffix = "M")) +
  scale_y_continuous(
    name = "Average User Rating",
    sec.axis = sec_axis(~ . * 30, name = "Total Watch Hours")
  ) +
  scale_color_manual(values = c("User Rating" = "red", "Watch Hours" = "blue")) +
  labs(title = "Production Budget Impact on User Ratings & Engagement",
       subtitle = "Relationship between investment, user satisfaction, and viewing hours",
       x = "Production Budget (Millions)",
       color = "Metric") +
  theme_minimal() +
  theme(legend.position = "bottom")

p3
```

#### Based on the Netflix Originals budget performance analysis:

No clear correlation between budget and ratings - High-budget films ($19M) show similar user ratings (3.0-3.8) as low-budget titles ($200K)

Cost efficiency varies widely - Some low-budget films achieve high engagement (129 watch hours for $220K) while expensive productions sometimes underperform

Optimal budget range emerges - Mid-budget films ($5-15M) often deliver balanced performance in both ratings and viewership hours

### 3. Recommendation Engine Effectiveness

####  Question: How effective are our recommendations? What's the click-through rate by recommendation type and position?

```{r}

recommendation_effectiveness <- recommendation_logs %>%
  group_by(recommendation_type, position_in_list) %>%
  summarize(
    total_recommendations = n(),
    clicks = sum(was_clicked),
    click_through_rate = clicks / total_recommendations * 100
  ) %>%
  arrange(position_in_list, desc(click_through_rate))

recommendation_effectiveness

```

3. Recommendation Engine Effectiveness

```{r}
# Focus on positions 1-5 only (usually most important)
rec_effectiveness_overall <- recommendation_logs %>%
  group_by(recommendation_type) %>%
  summarize(
    total_recommendations = n(),
    clicks = sum(was_clicked),
    click_through_rate = clicks / total_recommendations * 100,
    .groups = 'drop'
  )

# Plot overall performance by recommendation type
p4_overall <- ggplot(rec_effectiveness_overall, 
                     aes(x = reorder(recommendation_type, click_through_rate), 
                         y = click_through_rate, 
                         fill = recommendation_type)) +
  geom_col(alpha = 0.9, width = 0.7) +
  geom_text(aes(label = paste0(round(click_through_rate, 1), "%")), 
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(title = "OVERALL CLICK-THROUGH RATE BY RECOMMENDATION TYPE",
       subtitle = "Which recommendation strategies perform best overall",
       x = "Recommendation Type", 
       y = "Click-Through Rate (%)",
       fill = "Recommendation Type") +
  scale_fill_brewer(palette = "Pastel1") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, margin = margin(b = 10)),
    plot.background = element_rect(fill = "#faf9f7", color = NA),
    panel.background = element_rect(fill = "#fefefe", color = NA),
    panel.grid.major = element_line(color = "#f0f0f0"),
    panel.grid.minor = element_blank(),
    legend.position = "none"  # Remove legend since x-axis shows types
  )

p4_overall
```

####  Based on the overall recommendation engine performance:

  * New releases drive highest engagement - 15.55% CTR, suggesting users are most interested in fresh content
  * All algorithms perform similarly well - CTRs cluster tightly between 14.76-15.55%, indicating balanced strategy effectiveness
  * Genre-based recommendations lag slightly - 14.76% CTR, potentially due to less personalization than other methods
  * Minimal performance gaps - Only 0.79% difference between best and worst performing recommendation types


### 4. Geographical Content Preferences

####  Question: How do content preferences (genres, ratings) vary across different countries and age groups?

```{r}


geo_preferences <- watch_history %>%
  inner_join(users %>% select(user_id, country, age), by = "user_id") %>%
  inner_join(movies %>% select(movie_id, genre_primary, rating, imdb_rating), by = "movie_id") %>%
  group_by(country, age_group = cut(age, breaks = c(0, 25, 35, 50, 100))) %>%
  summarize(
    top_genre = names(which.max(table(genre_primary))),
    avg_imdb_rating = mean(imdb_rating, na.rm = TRUE),
    avg_user_rating = mean(user_rating, na.rm = TRUE),
    total_watch_hours = sum(watch_duration_minutes, na.rm = TRUE) / 60
  ) %>%
  arrange(country, age_group)

geo_preferences
```

#### Key Findings:

Adventure genre dominates universally - All age groups in both Canada and USA consistently prefer Adventure content over other genres
Minimal rating variation across demographics - IMDB ratings (6.07-6.09) and user ratings (3.29-3.42) show remarkably little difference between countries and age groups

Middle-aged users drive highest engagement - 35-50 age group watches the most content in both countries (12,578 hours in Canada, 28,625 hours in USA)

USA shows significantly higher consumption - American users watch 2-3x more content than Canadian users across all age segments


4. Geographical Content Preferences

```{r}
p7_top <- ggplot(top_genres, aes(x = age_group, y = percentage, fill = genre_primary)) +
  geom_col(alpha = 0.9, position = "dodge", width = 0.8) +
  facet_wrap(~country, ncol = 2) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_dodge(0.8), vjust = -0.5, size = 2.8) +
  labs(title = "TOP 3 GENRES BY AGE AND COUNTRY",
       subtitle = "Most popular genres in each demographic segment",
       x = "Age Group", y = "Percentage of Views", 
       fill = "Top Genres") +
  scale_fill_brewer(palette = "Set3") +  # Also try "Paired", "Set2", "Pastel1"
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, margin = margin(b = 15)),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

p7_top

```
####  Key findings:

Adventure genre consistently ranks #1 across all age groups in both Canada and USA (6.7-7.2% of views)

Genre preferences shift with age - Younger audiences prefer Animation and Sci-Fi, while older groups favor Action and War content

USA shows stronger genre concentration - Top genres capture higher percentage of total views compared to Canada

Older demographics (50+) show most distinct tastes - Different secondary genres emerge compared to younger age groups