---
title: "Project3_watch_history_data_cleaning"
Author: "Mehreen Ali Gillani"
output: html_document
date: "2025-10-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```
##  Data Cleaning of watch_history table of Netflix

### Step 1. Import Libraries

```{r import libraries}
library(tidyverse)
library(dplyr)
library(naniar)
library(gtExtras)
library(stringr)
library(mice)
library(lubridate)
```
### Step 2. Read watch_history dataset and store it in a Dataframe

```{r read csv}
watch_history_url = 'https://raw.githubusercontent.com/mehreengillani/DATA607_project3/refs/heads/main/watch_history.csv'
watch_history = read.csv(watch_history_url)
colnames(watch_history)

#data structure
str(watch_history)
```

### Step 3: Data Cleaning

####  3.1 Handling missing values
```{r identify missing values}

# Identify missing values
miss_var_summary(watch_history) %>%
  gt() %>%
  gt_theme_guardian() %>%
  tab_header(title="Missing data in watch_history table")

#plot
gg_miss_var(watch_history)
``` 

Immediate Assessment

High missingness (>79%): user_rating I will delete this column
for watch_duration_minutes, progress_percentage missing variable percentage is 11.7 and 8.11 so we will replace missing values with multiple imputation 
```{r}
summary(watch_history$user_rating)
```


```{r handling missing data}

# Keep all columns, only impute the specific missing ones

# Perform imputation on selected columns only
temp_impute <- watch_history %>%
  select(watch_duration_minutes, progress_percentage, device_type, action, quality)

imputed_temp <- mice(temp_impute, m = 1, method = "pmm", printFlag = FALSE)
imputed_temp_complete <- complete(imputed_temp)

# Create clean dataset keeping original user_rating as-is
watch_history_clean <- watch_history
watch_history_clean$watch_duration_minutes <- imputed_temp_complete$watch_duration_minutes
watch_history_clean$progress_percentage <- imputed_temp_complete$progress_percentage

# Check for missing data - user_rating should still have NAs
miss_var_summary(watch_history_clean)
```

Check density plot after imputation to see if our distribution is still the same 
```{r}
# Density plots of imdb_ratings before and after imputation
par(mfrow = c(1, 2))
plot(density(watch_history$watch_duration_minutes, na.rm = TRUE), main = "watch_duration_minutes Density \n before multiple imputation")
plot(density(watch_history_clean$watch_duration_minutes, na.rm = TRUE), main = "watch_duration_minutes \n Density after imputation")

plot(density(watch_history$progress_percentage, na.rm = TRUE), main = "progress_percentage Density \n before multiple imputation")
plot(density(watch_history_clean$progress_percentage, na.rm = TRUE), main = "progress_percentage \n Density after imputation")

par(mfrow = c(1, 1))
```
```{r}
summary(watch_history_clean$user_rating)
```

Distribution is almost the same after imputation.

Lets Calculate before/after statistics

```{r}

# Calculate before/after statistics of watch_duration_minutes
before <- watch_history$watch_duration_minutes[!is.na(watch_history$watch_duration_minutes)]
after <- watch_history_clean$watch_duration_minutes

# Basic statistics comparison
stats_before <- c(
  mean = mean(before),
  sd = sd(before),
  median = median(before),
  min = min(before),
  max = max(before),
  n = length(before)
)

stats_after <- c(
  mean = mean(after),
  sd = sd(after),
  median = median(after),
  min = min(after),
  max = max(after),
  n = length(after)
)

comparison <- data.frame(Before = stats_before, After = stats_after, Change = stats_after - stats_before, percentage = ((stats_after - stats_before)/stats_before) * 100)
print(round(comparison, 3))
```

Watch Duration Minutes - Imputation Impact Summary

✅ Minimal Impact - Excellent Results

Mean: -0.471 minutes (-0.72%)

Negligible decrease in average watch time
Standard Deviation: -1.551 minutes (-2.27%)

Slight reduction in variability, actually beneficial
Median: +0.100 minutes (+0.20%)

Virtually unchanged central tendency
Range: Perfectly preserved

Min (0.2 min) and Max (799.3 min) unchanged
No artificial constraints introduced
Sample Size: +12,332 records (+13.3%)

Successfully recovered missing data points

```{r}

# Calculate before/after statistics of progress_percentage
before <- watch_history$progress_percentage[!is.na(watch_history$progress_percentage)]
after <- watch_history_clean$progress_percentage

# Basic statistics comparison
stats_before <- c(
  mean = mean(before),
  sd = sd(before),
  median = median(before),
  min = min(before),
  max = max(before),
  n = length(before)
)

stats_after <- c(
  mean = mean(after),
  sd = sd(after),
  median = median(after),
  min = min(after),
  max = max(after),
  n = length(after)
)

comparison <- data.frame(Before = stats_before, After = stats_after, Change = stats_after - stats_before, percentage = ((stats_after - stats_before)/stats_before) * 100)
print(round(comparison, 3))
```

Progress Percentage - Imputation Impact Summary

✅ Exceptional Preservation - Nearly Perfect

Mean: -0.121% (-0.24%)

Virtually identical average completion rate
Standard Deviation: -0.081% (-0.28%)

Minimal change in variability
Median: -0.100% (-0.20%)

Negligible shift in central tendency
Range: Perfectly preserved

Min (0%) and Max (100%) unchanged
Full scale integrity maintained
Sample Size: +8,514 records (+8.82%)

Significant recovery of missing progress data


```{r}
#check categorical values and see their unique values
unique(watch_history_clean$action)
unique(watch_history_clean$quality)
```
Replace NA with unknown 
```{r}
watch_history_clean1 <- watch_history_clean %>%
  mutate(
    # Action: convert NA to "unknown" ← THIS LINE HANDLES ACTION NA
    action = as.factor(ifelse(is.na(action), "unknown", as.character(action))),
    
    # Quality: preserve ordering, add "unknown" as lowest level
    quality = case_when(
      is.na(quality) ~ "unknown",
      TRUE ~ as.character(quality)
    ),
    quality = factor(quality, levels = c("unknown", "SD", "HD", "4K", "HDR"), ordered = TRUE)
  )
unique(watch_history_clean$action)
unique(watch_history_clean$quality)
```

#### Step 3.2: Removing Duplicates

```{r check duplicates}
# Identify total duplicates
sum(duplicated(watch_history_clean1))
duplicated_percent <- (sum(duplicated(watch_history_clean1)) / nrow(watch_history_clean1) * 100)
print(duplicated_percent)

```

The percentage of duplicates is 4.07 so we will delete duplicated rows

```{r remove duplicates}

# Remove ALL duplicate rows (keeping first occurrence)
watch_history_unique <- distinct(watch_history_clean1)

# Verify removal
print(paste("Before:", nrow(watch_history_clean1), "rows"))
print(paste("After:", nrow(watch_history_unique), "rows"))
print(paste("Removed:", nrow(watch_history_clean1) - nrow(watch_history_unique), "duplicate rows"))
```

####  Step: 3.3 Data Type Conversion

```{r check data types}
# Basic structure
str(watch_history_unique)
```
```{r}
unique(watch_history_unique$quality)
```

```{r data type conversion}


watch_history_final <- watch_history_unique %>%
  mutate(
    # Date conversion
    watch_date = as.Date(watch_date),
    
    # Logical conversion
    is_download = as.logical(is_download),
    
    # Ordered factors with logical hierarchies
    device_type = factor(device_type, 
                        levels = c("Mobile", "Tablet", "Laptop", "Desktop", "Smart TV"),
                        ordered = TRUE),
    
    # FIXED: Removed trailing comma after "completed"
    action = factor(action,
                   levels = c("unknown", "stopped", "paused", "started", "completed"),
                   ordered = TRUE),
    
    quality = factor(quality,
                    levels = c("unknown", "SD", "HD", "4K", "HDR"),
                    ordered = TRUE),
    
    # Regular factors
    location_country = as.factor(location_country)
  )
  
```

We have converted watch_date to date format, is_download logical and
categorical values to factors:
device_type, action, genre_secondary, quality, location_country

```{r}
#check new structure 
str(watch_history_final)
```

####  Step 3.4. Handling Outliers

```{r fine numeric columns}
# Find numeric columns
numeric_cols <- names(watch_history_final)[sapply(watch_history_final, is.numeric)]

cat("Numeric columns:\n")
print(numeric_cols)

```


BoxPlot before Outlier handling 

```{r before outlier detection}

boxplot(watch_history_final$watch_duration_minutes, main = "watch_duration_minutes Outliers", col = "lightblue")
boxplot(watch_history_final$progress_percentage, main = "progress_percentage Outliers", col = "lightgreen")

```
From these plots we can see there are no outliers for progress_percentage. But there are in watch_duration_minutes

Create new column duration_category in dataset

```{r}
watch_history_final <- watch_history_final %>%
  mutate(
    duration_category = factor(
      case_when(
        watch_duration_minutes == 0 ~ "0 minutes",
        watch_duration_minutes <= 5 ~ "1-5 minutes",
        watch_duration_minutes <= 30 ~ "6-30 minutes",
        watch_duration_minutes <= 60 ~ "31-60 minutes",
        watch_duration_minutes <= 120 ~ "1-2 hours",
        watch_duration_minutes <= 240 ~ "2-4 hours", 
        watch_duration_minutes <= 480 ~ "4-8 hours",
        TRUE ~ "Other"
      ),
      levels = c("0 minutes", "1-5 minutes", "6-30 minutes", "31-60 minutes", 
                 "1-2 hours", "2-4 hours", "4-8 hours", "Other"),
      ordered = TRUE
    )
  )

```

```{r}
# Calculate outlier bounds using IQR method
duration_data <- watch_history_final$watch_duration_minutes

Q1 <- quantile(duration_data, 0.25)
Q3 <- quantile(duration_data, 0.75)
IQR <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

cat("=== OUTLIER DETECTION ===\n")
cat("Q1 (25th percentile):", Q1, "minutes\n")
cat("Q3 (75th percentile):", Q3, "minutes\n") 
cat("IQR:", IQR, "minutes\n")
cat("Lower bound:", lower_bound, "minutes\n")
cat("Upper bound:", upper_bound, "minutes\n")

outliers <- sum(duration_data > upper_bound)
cat("Outliers detected:", outliers, "(", round(outliers/length(duration_data)*100, 2), "%)\n")


```
Analysis of Results

Key Issues Identified:
❌ Lower bound is negative (-50.5 minutes) - impossible for watch duration
❌ Upper bound seems too low (161.5 minutes = 2.7 hours)
✅ Only 3.83% outliers - manageable amount

Set lower value to 0 and upper value to 8 hours

```{r}

# Set the bounds
LOWER_BOUND <- 0      # Minimum 0 minutes
UPPER_BOUND <- 8 * 60 # Maximum 8 hours (480 minutes)

# Apply the capping
watch_history_final <- watch_history_final %>%
  mutate(
    watch_duration_minutes = case_when(
      watch_duration_minutes < LOWER_BOUND ~ LOWER_BOUND,
      watch_duration_minutes > UPPER_BOUND ~ UPPER_BOUND,
      TRUE ~ watch_duration_minutes
    )
  )

# Calculate statistics
original_data <- watch_history_final$watch_duration_minutes  # Before any changes if you saved it
capped_below <- sum(watch_history_final$watch_duration_minutes < LOWER_BOUND, na.rm = TRUE)
capped_above <- sum(watch_history_final$watch_duration_minutes > UPPER_BOUND, na.rm = TRUE)

cat("=== BOUNDS CAPPING RESULTS ===\n")
cat("Lower bound:", LOWER_BOUND, "minutes\n")
cat("Upper bound:", UPPER_BOUND, "minutes (", UPPER_BOUND/60, "hours)\n")
cat("Values capped below:", capped_below, "\n")
cat("Values capped above:", capped_above, "\n")
cat("New range: [", min(watch_history_final$watch_duration_minutes), ", ", 
    max(watch_history_final$watch_duration_minutes), "] minutes\n", sep = "")
```



BoxPlot after outlier handling with cap method

```{r after outlier cap}

boxplot(watch_history_final$watch_duration_minutes, main = "watch_duration_minutes after applying cap", col = "lightblue")

# Reset plot layout
par(mfrow = c(1, 1))
```

```{r}

# Simple bar plot
ggplot(watch_history_final, aes(x = duration_category)) +
  geom_bar(fill = "steelblue", alpha = 0.7) +
  labs(title = "Watch Duration Categories",
       x = "Duration Category", 
       y = "Number of Sessions") +
  theme_minimal()
```


####  3.5 String Cleaning


```{r find textCols_string_cleaning}
# Find all text columns
text_cols <- names(watch_history_final)[sapply(watch_history_final, function(x) is.character(x) | is.factor(x))]

cat("Text columns found:\n")
print(text_cols)
```

Text cleaning: convert to lower case, convert empty string, remove whitespace

```{r text_cleaning_p2}
# Perform comprehensive text cleaning
watch_history_cleaned <- watch_history_final %>%
  mutate(across(all_of(text_cols), ~ {
    if(is.character(.)) {
      # Remove leading/trailing whitespace
      cleaned <- trimws(.)
      # Convert to lowercase for standardization
      cleaned <- tolower(cleaned)
      # Remove multiple spaces
      cleaned <- gsub("\\s+", " ", cleaned)
      # Convert empty strings to NA
      cleaned <- ifelse(cleaned == "", NA, cleaned)
      return(cleaned)
    } else {
      return(.)
    }
  }))
```


```{r}
colnames(watch_history_cleaned)

unique(watch_history_cleaned$quality)
```

####  3.6 Data Validation

```{r Validate variables}

# data validation function

validate_cleaned_data <- function(df) {
  cat("=== DATA VALIDATION REPORT ===\n\n")
  
  # 1. Basic info
  cat("1. BASIC INFO\n")
  cat("   Rows:", nrow(df), "\n")
  cat("   Columns:", ncol(df), "\n")
  cat("   Column names:", paste(names(df), collapse = ", "), "\n\n")
  
  # 2. Missing values
  cat("2. MISSING VALUES\n")
  missing_count <- sapply(df, function(x) sum(is.na(x)))
  for(col in names(missing_count)) {
    if(missing_count[col] > 0) {
      cat("   ", col, ":", missing_count[col], "missing values\n")
    }
  }
  if(all(missing_count == 0)) cat("   No missing values found\n")
  cat("\n")
  
  # 3. Data types - FIXED
  cat("3. DATA TYPES\n")
  for(col in names(df)) {
    cat("   ", col, ":", class(df[[col]])[1], "\n")
  }
  cat("\n")
  
  # 4. Basic value checks
  cat("4. VALUE CHECKS\n")
  
  # Check numeric columns
  numeric_cols <- names(df)[sapply(df, is.numeric)]
  for(col in numeric_cols) {
    cat("   ", col, ":\n")
    cat("      Min:", min(df[[col]], na.rm = TRUE), "\n")
    cat("      Max:", max(df[[col]], na.rm = TRUE), "\n")
    cat("      NA's:", sum(is.na(df[[col]])), "\n")
  }
  
  # Check categorical columns
  cat_cols <- names(df)[sapply(df, is.character) | sapply(df, is.factor)]
  for(col in cat_cols) {
    unique_vals <- length(unique(df[[col]]))
    cat("   ", col, ":", unique_vals, "unique categories\n")
  }
  
  return(TRUE)
}

# Run the validation
validate_cleaned_data(watch_history_cleaned)

```

Although user_rating has 80468 missing values. we are keeping this column because it is important for data analysis.
This validation step is crucial before any analysis to ensure your results are reliable and meaningful!

Write cleaned data to another csv file:

```{r}
# Write to CSV file
write.csv(watch_history_cleaned, "watch_history_clean.csv", row.names = FALSE)

cat("Cleaned data saved as 'watch_history_clean.csv'\n")
cat("File location:", getwd(), "\n")

```

```{r check new file}
watch_history_cleaned_data <- read_csv('watch_history_clean.csv')
#str(watch_history_cleaned_data)

```

